Value construction through sequential sampling explains serial dependencies in decision making

Version 1

Abstract

Abstract

Many decisions are expressed as a preference for one item over another. When these items are familiar, it is often assumed that the decision maker assigns a value to each of the items and chooses the item with the highest value. These values may be imperfectly recalled, but are assumed to be stable over the course of an interview or psychological experiment. Choices that are inconsistent with a stated valuation are thought to occur because of unspecified noise that corrupts the neural representation of value. Assuming that the noise is uncorrelated over time, the pattern of choices and response times in value-based decisions are modeled within the framework of Bounded Evidence Accumulation (BEA), similar to that used in perceptual decision-making. In BEA, noisy evidence samples accumulate over time until the accumulated evidence for one of the options reaches a threshold. Here, we argue that the assumption of temporally uncorrelated noise, while reasonable for perceptual decisions, is not reasonable for value-based decisions. Subjective values depend on the internal state of the decision maker, including their desires, needs, priorities, attentional state, and goals, which may change over time. These internal states may change over time, or undergo revaluation, as will the subjective values. We reasoned that these hypothetical value changes should be detectable in the pattern of choices made over a sequence of decisions. We reanalyzed data from a well-studied task in which participants were presented with pairs of snacks and asked to choose the one they preferred. Using a novel algorithm (Reval), we show that the subjective value of the items changes significantly during a short experimental session (about 1 hour). Values derived with Reval explain choice and response time better than explicitly stated values. They also better explain the BOLD signal in the ventromedial prefrontal cortex, known to represent the value of decision alternatives. Revaluation is also observed in a BEA model in which successive evidence samples are not assumed to be independent. We argue that revaluation is a consequence of the process by which values are constructed during deliberation to resolve preference choices.

Introduction

A central idea in decision theory and economics is that each good can be assigned a scalar utility value that reflects its desirability. The concept of utility, or subjective value, provides a common currency for comparing dissimilar goods (e.g., pears and apples) such that decision-making can be reduced to estimating the utility of each good and comparing them (von Neumann and Morgenstern, 1944; Samuelson, 1937; Montague and Berns, 2002). The idea is supported by studies that have identified neurons that correlate with the subjective value of alternatives in various brain structures, most notably the ventromedial prefrontal cortex, and it is so pervasive that decisions based on preferences are often referred to as “value-based decisions” (Kable and Glimcher, 2007; Kim et al., 2008; Padoa-Schioppa and Assad, 2006).

Choice and response time (RT) in simple perceptual and mnemonic decisions are often modeled within the framework of bounded evidence accumulation (BEA). The framework posits that evidence samples for and against the different options are accumulated over time until the accumulated evidence for one of the options reaches a threshold or bound (Ratcliff, 1978; Gold and Shadlen, 2007). A case in point is the random dot motion (RDM) discrimination task, in which participants must decide whether randomly moving dots have net rightward or leftward motion, while the experimenter controls the proportion of dots moving coherently in one direction, termed the motion strength (e.g., Gold and Shadlen, 2007). BEA models explain the choice, RT, and confidence in the RDM task under the assumption that the rate of accumulation, often termed the drift rate, depends on motion strength (van Den Berg et al., 2016; Kiani et al., 2014). Value-based decisions have also been modeled within the framework of BEA. The key assumption is that at any given time, decision-makers only have access to a noisy representation of the subjective value of each item, and the drift rate depends on the difference between the subjective values of the items (Krajbich et al., 2010; Thomas et al., 2019; Sepulveda et al., 2020; Bakkour et al., 2019).

A condition that renders the BEA framework normative is that the noise corrupting the evidence samples is independent, or equivalently, that the evidence samples are conditionally independent given the drift rate. For example, in modeling the RDM and other perceptual decision making tasks, evidence samples are assumed to be independent of each other, conditioned on motion strength and direction (e.g., Zylberberg et al., 2016). This assumption is sensible because (i) the main source of stochasticity in perceptual decision making is the noise affecting the sensory representation of the evidence, which has a short-lived autocorrelation, and (ii) these decisions are often based on an evidence stream (e.g., a dynamic random dot display) that provides conditionally independent samples, by design. The assumption of conditional independence justifies the process of evidence accumulation, because accumulation (or averaging) can only remove the noise components that are not shared by the evidence samples.

For value-based decisions, the assumption of conditional independence is questionable. Alternatives often differ across multiple attributes (e.g., Busemeyer and Townsend, 1993; Tversky, 1977). For example, when choosing between different snacks, they may differ in calories, healthiness, palatability, and so on (Suzuki et al., 2017). The weight given to each attribute depends on the decision-maker’s internal state (Noguchi and Stewart, 2018; Juechems and Summerfield, 2019). This internal state includes desires, needs, priorities, attentional state and goals. We use the term mindset, or state of mind, to refer to all of these internal influences on valuation. A mindset can be persistent. For example, a famished decision-maker may prioritize the nutritional content of each food when making a choice. Under less pressing circumstances, the salience of an attribute may be suggested by snack alternatives themselves. For example, seeing French fries may make us aware that we crave something salty, and saltiness becomes a relevant attribute informing the current decision and possibly future decisions too. The examples illustrate how a decision-maker’s mindset can shift rapidly or meander, based on the attributes in focus or the identity of the items under consideration (Shadlen and Shohamy, 2016; Stewart et al., 2006). Importantly, mindset is dynamic. It can change abruptly, motivated by a thought in an earlier trial or by interoception during deliberation (e.g., thirst). Unlike perceptual decision-making, where the expectation of a sample of evidence is thought to be fixed, conditional on the stimulus, the expectation of the evidence bearing on preference is itself potentially dynamic.

We sought to test the notion that the desirability of an item changes as a result of the deliberation that leads to a choice. We hypothesized that if subjective values are dynamic, then value-based decisions should exhibit serial dependencies when multiple decisions are made in a sequence. A choice provides information not only about which option is preferred, but also about the decision maker’s mindset at the moment of the choice (e.g., whether they prioritize satiation or palatability). Therefore, a choice is informative about future choices because the decision maker’s mindset is likely to endure longer than a single decision, or even multiple decisions.

We reanalyzed data from Bakkour et al. (2019). Participants were presented with pairs of snacks and had to choose the one they preferred. This Food choice task has been used extensively to study the sequential sampling process underlying value-based decisions (e.g., Krajbich et al., 2010). Crucially, in the Bakkour et al. (2019) experiment, each item was presented multiple times, allowing us to infer how preference for an item changes during a single experimental session. Using a novel algorithm we call Reval, we show that the subjective value of items changed over the session. The revaluation was replicated in a sequential sampling model in which successive samples of evidence are not assumed to be conditionally independent. We argue that the revaluation process we observed reflects a process by which the value of the alternatives is constructed during deliberation by querying memory and prospecting for evidence that bears on desirability (Lichtenstein and Slovic, 2006; Johnson et al., 2007).

Results

Food choice task

We re-examined data from a previous study in which 30 participants completed a food choice task (Bakkour et al., 2019). In each trial, participants were shown a pair of snack images and had to choose which one they would prefer to consume at the end of the study (Fig. 1A). Prior to the main experiment (conducted in an MRI scanner), participants were asked to indicate their willingness to pay for each snack item on a scale from 0 to US$3 (Fig. 1B). We refer to these explicitly reported values as e-values, or ve.

Food choice task.

(A) In the main task, participants were presented with pairs of snack items and asked to choose which one they would prefer to consume at the end of the session. After making their choice, the chosen item was highlighted by a square box for an additional 0.5 s. Each of the 30 participants completed 210 trials, with each item appearing 7 times during the experiment. A subset of 60 item pairs were repeated once.

(B) In an initial ‘ratings’ task, participants were shown 60 individual appetizing snack items and asked to indicate how much they would be willing to pay for each item using a monetary scale ranging from $0 to $3.

(C) Proportion of trials in which participants selected the right item as a function of the difference in value between the right and left items (Δve). Proportions were first determined for each participant and then averaged across participants. Error bars indicate the standard error of the mean (s.e.m.) across participants.

(D) Mean response time as a function of the difference in value between the right and left items. Error bars indicate the s.e.m. across participants. Red curves in panels C-D are fits of a drift-diffusion model (DDM).

The data from Bakkour et al. (2019) replicate the behavior typically observed in the task. Both choice and response time were systematically related to the difference in e-value, (Δve), between the right and left items. Participants were more likely to choose the item to which they assigned a higher value during the rating phase (p<0.0001; H0 ∶ β1 = 0; Eq. 2). They were also more likely to respond faster when the absolute value of the difference between the items was greater (p<0.0001; H0 ∶ β1 = 0; Eq. 3).

The relationship between Δve, choice, and response time is well described by a bounded evidence accumulation model (Krajbich et al., 2010; Bakkour et al., 2019). The solid lines in Fig. 1C-D illustrate the fit of such a model in which the drift rate depends on Δve. Overall, the behavior of our participants in the task is similar to that observed in other studies using the same task (e.g., Krajbich et al., 2010; Folke et al., 2016; Sepulveda et al., 2020).

Limited power of explicit reports of value to explain binary choices

An intriguing aspect of the decision process in the food choice task is its highly stochastic nature. This is evident from the shallowness of the choice function (Fig. 1C): participants chose the item with a higher e-value in only 64% of the trials. This variability is typically attributed to unspecified noise when recalling item values from memory (e.g., Krajbich et al., 2010). An alternative explanation is rooted in constructive value theories, which suggest that the value of each item is constructed, not retrieved, during the decision process (Lichtenstein and Slovic, 2006; Shadlen and Shohamy, 2016; Johnson et al., 2007). This construction process is sensitive to the context in which it is elicited (e.g., the identity of items being compared), so the values reported during the valuation process may differ from those used in the choice task. According to this idea, the apparently stochastic choice is a veridical reflection of the constructed values.

If this were true, then the choice on any one cynosure trial—that is, the trial we are scrutinizing—would be better explained by values inferred from the choices on the other trials than by the e-values. We therefore compared two regression models that produce the log odds of the choice on each cynosure trial. The first regression model uses the e-values plus a potential bias for the left or right item. The second regression model includes one regression coefficient per item plus a left/right bias. It uses all the other trials (except repetitions of the identical pair of items) to establish the weights. While this model has more free parameters, the comparison is valid because we are using the models to predict the choices made on trials that were not used for model fitting. The better model is the one that produces larger log odds of the choice on the cynosure trial. As shown in Fig. 2, the second regression model is superior.

Individual choices are better explained by values inferred from the other trials than values reported in the ratings task.

Gray data points represent the total log-likelihood of each participant’s choices, given two types of predictions: (abscissa) from a logistic regression, fit to the explicit values; (ordinate) from a procedure that infers the values based on choices on the other trials. Predictions derived from the other trials is better in all but four participants. The red markers were obtained using the same procedure, applied to choices simulated under the assumption that the e-values are the true values of the items. It shows that the inferential procedure is not guaranteed to improve predictions.

To ensure that this result is not produced artifactually from the algorithm, we performed the same analysis on simulated data. We fit the experimentally observed choices using a logistic regression model with Δve and an offset as independent variables, and simulated the choices by sampling from Bernoulli distributions with parameter, p, specified by the logistic function that best fit each participant’s choices (i.e., weighted-coin flips). We repeated the model comparison using the simulated choices and found that, contrary to what we observed in the experimental data, the model using explicit value reports is the better predictor (Fig. 2, red).

Taken together, these analyses show that explicit value reports have limited power to predict choices, which partially explains their apparent stochasticity. In the following sections, we elaborate on this observation. Not only do the values used to make the binary choices differ from the e-values, they drift apart during the experiment. We show that these changes arise through the deliberative process leading to the preference decisions themselves.

Preferences change over the course of the experiment

In the experiment, a subset of the 60 snack pairs were presented twice, in a random order within the sequence of trials. These trials allow us to assess whether preferences change over the course of a session. For these duplicated item pairs, we calculate the average number of times that the same item was chosen on both presentations—which we refer to as the match probability. Participants were more likely to select the same option when presentations of the same pair were closer in time (Fig. 3). To assess the significance of this effect, we fit a logistic regression model using all pairs of trials with identical stimuli to predict the probability that the same item would be chosen on both occasions. The regression coefficient associated with the number of trials between repetitions was negative and highly significant (p<0.0001; t-test, Eq. 8). It therefore follows that preferences are not fixed, not even over the course of a single experimental session.

Preferences change over time.

Probability of making the same choice on the two trials with the same item pair, shown as a function of the difference in trial number between them (Δtr). Trial pairs with identical items (N=1726) were sorted by Δtr, and the match probabilities were smoothed with a boxcar function with a width of 100 observations.

Choice alternatives undergo revaluation

We propose a simple algorithm to characterize how preferences changed over the course of the session. It assumes that on each decision, the value of the chosen item increases by an amount equal to δ, and the value of the unchosen item decreases by the same amount (Fig. 4A). We refer to the updated values as r-values, or vr, as opposed to the explicitly reported values (e-values).

Revaluation algorithm.

(A) Schematic example of the revaluation algorithm applied to one decision. After a choice between items A and B, the value of the chosen item is increased by δ and the value of the unchosen item is decreased by the same amount. (B) Example of value changes due to revaluation, for three items, as a function of the presentation number within the session. In the experiment, each item was presented 7 times. (C) For a representative participant, deviance of the logistic regression model that uses the revalued values to explain the choices, for different values of δ. The best fitting value is ∼$0.15. The inset shows a histogram of the best-fitting δ values across participants.

Fig. 4B illustrates how the value of the items changes over the course of the session, for a given value of δ, for three snack items. For example, while the item shown with the green curve is initially very valuable, as indicated by its high initial rating, its value decreases over the course of the session each time it was not selected.

We determined the degree of revaluation that best explained the participants’ choices. For each participant, we find the value of δ that minimizes the deviance of a logistic regression model that uses the r-values to fit the choices made on each trial, where pchoice is the probability of choosing the item that was presented on the right. The r-values are initialized to the explicitly reported values for all items, and they are updated by plus or minus δ when an item is chosen or rejected, respectively. Importantly, the updated values only affect future decisions involving the items.

Fig. 4C shows the deviance of the logistic regression model for a representative participant, as a function of δ. For this participant, the best explanation of the choices is obtained with a value of δ ≈ $0.15. We fit the value of δ independently for each participant to minimize the deviance of the logistic regression model fit to the choices. On average, each choice changed the value of the chosen and unchosen items by $0.18 ± 0.016 (mean ± s.e.m., Fig. 4C, inset).

The values derived from the Reval algorithm explain the choices better than the explicit value reports. The choices are more sensitive to variation in Δvr, evidenced by the steeper slope (Fig. 5A). Further, when Δvr and Δve are allowed to compete for the same binomial variance, the former explains away the latter. This assertion is supported by a logistic regression model that incorporates both Δve and Δvr as explanatory variables (Eq. 7). The coefficient associated with Δve is not significantly different from zero (p = 0.32, t-test; Eq. 7) while the one associated with Δvr remains positive and highly significant (p<0.0001).

Revaluation explains choice and RT better than explicit values.

(A) Proportion of rightward choices (top) and mean response time (bottom) as function of the difference in r-value between the two items. The red solid lines are fits of a drift-diffusion model that uses the r-values. The dashed line corresponds to the fits of a DDM that uses the e-values (same as in Fig. 1C-D). Error bars indicate s.e.m. across trials. Participants are more sensitive to r-values than e-values (top) and the r-values better explain the full range of RTs (bottom). (B) Model-free, empirical support for the superiority of r-values in individual subjects. Data points represent the difference in mean RTs between difficult and easy decisions. Positive values indicate that difficult decisions take longer on average than easy ones. Difficult and easy are defined relative to the median of the absolute value of Δve (left) or Δvr (right). The lines connect the mean RTs of each participant. P-value is from a paired t-test.

More surprisingly, Reval allows us to explain the response times better than the explicit value reports, even though RTs were not used to establish the r-values. We used the r-values to fit a drift-diffusion model to the single-trial choice and response time data, and compared this model with the one that was fit using the e-values (Fig. 5A). To calculate the fraction of RT variance explained by each model, we subtracted from each trial’s RT the models’ expectation, conditional on Δvx (with x ∈ {e, r}) and choice. The model that relies on the r-values explains a larger fraction of variance in RT than the model that relies on the e-values (Fig. S1). This indicates that the re-assignment of values following Reval improved the capacity of a DDM to explain the response times.

The application of Reval revealed that some decisions that were initially considered difficult, because Δve was small, were actually easy, because Δvr was large, and vice versa. Grouping trials by the Δvr led to a wider range of mean RTs compared to when we grouped them by Δve (Fig. 5A). The effect can also be observed for individual participants. For each participant, we grouped trials into two categories depending on whether the difference in value was less than or greater than the median difference. We then calculated the mean RT for each of the two groups of trials. The difference in RT between the two groups was greater when we grouped the trials using the r-values than when we used the e-values. This implies the r-values were better than the e-values at assessing the difficulty of a decision as reflected in the response time.

We verified that the improvement in fit was not just due to the additional free parameter (δ). To do this, we again used simulated choices sampled from logistic regression models fit to the participants’ choices, as we did for Fig. 2. Because the choices are sampled from logistic functions fit to the choice data, they lead to a psychometric function that is similar to that obtained with the experimental data. We reasoned that if revaluation were an artifact of the analysis method, then applying the revaluation algorithm to these simulated data should lead to values of δ and goodness of fit similar to those of the real data. To the contrary, the optimal values of δ for the simulated data were close to zero (Fig. 6A), and we found no difference in the RT median splits between e-values and r-values (Fig. 6B). This shows that the improvements in fit quality due to Reval are neither guaranteed nor an artifact of the procedure.

No revaluation in simulated data.

(A) Histogram of the best-fitting revaluation update (δ) for data simulated by sampling choices from a logistic function fit to the participants’ choices. The best-fitting δ values for the simulated choices are centered around 0. For reference, we have also included a histogram of the δ values obtained from the fits to the participants’ data, showing all positive values (gray).

(B) Similar to Fig. 5B, for the simulated data. The values obtained from Reval were no better than the explicit values at explaining the RTs, as expected, since the δ values were ∼0 and thus vr ≈ ve.

Imperfect value reports do not explain revaluation away

The idea that a choice can induce a change in preference is certainly not new (Festinger, 1957). Choice-induced preference change (CIPC) has been documented using a free-choice paradigm (Brehm, 1956), whereby participants first rate several items, and then choose between pairs of items to which they have assigned the same rating, and finally rate the items again. A robust finding is that items that were chosen are given higher ratings and items that were not chosen are given lower ratings relative to pre-choice ratings, leading to the interpretation that the act of choosing changes the preferences for the items under comparison. However, it has been suggested that the CIPC demonstrated with the free-choice paradigm can be explained as an artifact (Chen and Risen, 2010). Put simply, the initial report of value may be a noisy rendering of the true latent value of the item. If two items, A and B, received the same rating but A was chosen over B, then it is likely that the true value for item A is greater than for item B, not because the act of choosing changes preferences, but because the choices are informative about the true values of the items, which are unchanging.

We examined whether Reval could be explained by the same artifact. We considered the possibility that the items’ valuation in the choice phase are static but potentially different from those reported in the ratings phase. If the values are static, but different from those explicitly reported, then Reval could still improve choice and RT predictions by revealing the true subjective value of the items.

We reasoned that if values were static, the improvements we observed in the logistic fits when we applied Reval should be the same regardless of how we ordered the trials before applying it. To test this, we applied Reval in the direction in which the trials were presented in the experiment, and also in the reverse direction (i.e., from the last trial to the first). If the values were static, then the quality of the fits should be statistically identical in both cases. In contrast, we observed that the variance explained by Reval was greater (i.e., the deviance was lower) when it was applied in the correct order than when it was applied in the opposite order (Fig. 7; p<0.0001, paired t-test). This rules out the possibility that the values were static.

Reval is sensitive to trial order.

Deviance obtained by applying Reval to the trials in the order in which they were completed (abscissa) and in the reverse order (ordinate). Each point corresponds to a different participant. The deviance is greater (i.e., the fits are worse) when Reval is applied in the reverse direction.

Asymmetric value-updating for chosen and unchosen options

So far we have assumed that a choice increases the value of the chosen option by δ and decreases the value of the unchosen option by the same amount. Here, we evaluate the possibility that the degree of revaluation is different for the chosen and unchosen options. We fit a variant of the Reval algorithm with two values of δ, one for the chosen option (δchosen) and one for the unchosen option (δunchosen). Fig. 8 shows the values that best fit the data. Each point corresponds to one participant. It can be seen that the degree of revaluation is greater for the chosen option than for the unchosen option. As we speculate in the discussion, this result may be related to the unequal distribution of attention between the chosen and unchosen items (Krajbich et al., 2010).

Stronger revaluation for the chosen than for the unchosen item.

We fit a variant of the Reval algorithm that includes separate update values (δs) for the chosen and unchosen options. The best-fitting δ value for the chosen option (abscissa) is plotted against the best-fitting value for the unchosen option (ordinate). Each data point corresponds to one participant. The increase in value for the chosen option is greater than the decrease in value for the unchosen option (paired t-test).

Representation of revalued values in the ventromedial prefrontal cortex

Several brain areas, in particular the ventromedial prefrontal cortex (vmPFC), have been shown to represent the value of decision alternatives during value-based decisions (Kennerley et al., 2009; Plassmann et al., 2007; Bartra et al., 2013). Based on our finding that the r-values provide a better explanation of the behavioral data than the e-values, we reasoned that the r-values might explain the BOLD activity in these areas beyond that explained by the e-values. We included both the e-value and the r-value of the chosen item in a whole-brain regression analysis of BOLD activity. This parameterization reveals significant correlation of the BOLD signal in the vmPFC only with the r-value (Fig. 9 and Table S1), providing additional evidence for revaluation, as capturing a meaningful aspect of the data, in the sense that it accounts for the activity of brain areas known to reflect the value of the choice alternatives.

Revaluation reflected in BOLD activity in ventromedial prefrontal cortex.

Brain-wide fMRI analysis revealed a significant correlation between r-values and activity in the vmPFC, after controlling for e-values. Coordinates are reported in standard MNI space. Heatmap color bars range from z-stat = 2.3 to 3.2. Map was cluster corrected for familywise error rate at whole brain level (p < 0.05).

Revaluation in other datasets of the food-choice task

To assess the generality of our behavioral results, we applied Reval to other publicly available datasets. All involve binary choices between food snacks, similar to Bakkour et al. (2019). We analyze data from experiments reported in Folke et al. (2016) and from the two value-based decision tasks reported in Sepulveda et al. (2020).

In all cases, Reval yields results similar to those observed in the data from Bakkour et al. (2019). The values derived from Reval led to a better classification of choice difficulty than the explicit value reports (Fig. 10). These results show the generality of the revaluation process and allow us to rule out the possibility that the findings are specific to a particular dataset or laboratory.

Similar results observed in other datasets.

We applied the Reval method to other publicly available datasets of the food choice task. We use the same ΔV comparison as in Figs. 5 and 7. (A) Data from the experiment of Folke et al. (2016). Participants (N=28) reported their willingness to pay (WTP) for each of 16 common snack items. In the choice task, they were presented with each unique pair of items and asked to choose the preferred item. Each unique pair was presented twice for a total of 240 trials per participant. We use the same ΔV comparison as in Figs. 5 and 7 to assess whether r-values better explain RTs of the study participants. (B and C) Data from the experiment of Sepulveda et al. (2020). Participants (N=31) reported their willingness to pay (WTP) for each of 60 snack items. They were then presented with pairs of items to choose from. Pairs were selected based on participants’ WTP reports to provide comparisons between pairs of high-value, low-value and mixed-value items. The choice task was performed under two framing conditions: like-framing, select the more preferred item, and dislike framing, select the less preferred item. The task consisted of six alternating blocks of like- and dislike-framing (40 trials per block).

Is revaluation a byproduct of deliberation?

We hypothesize that the sequential dependencies we identified with Reval may be a corollary of the process by which values are constructed during deliberation. The subjective value of an item depends on the decision-maker’s mindset, which may change more slowly than the rate of trial presentations. Therefore, the subjective value of an item on a given trial may be informative about the value of the item the next time it is presented. Subjective values are not directly observable, but choices are informative about the items’ value.

We assessed the plausibility of this hypothesis with a bounded evidence accumulation model that includes a parameter that controls the correlation between successive evidence samples for a given item. We call this the correlated-evidence drift-diffusion model (ceDDM). We assume that the decision is resolved by accumulating evidence for and against the different alternatives until a decision threshold is crossed.

The model differs from standard drift-diffusion, where the momentary evidence is a sample drawn from a Normal distribution with expectation equal to Δve plus unbiased noise, . Instead, the value of each of the items evolves separately such that the expectations of its value updates are constructed as a Markov chain Monte Carlo (MCMC) process thereby introducing autocorrelation between successive samples of the unbiased noise (see Methods). Crucially, the correlation is not limited to the duration of a trial but extends across trials containing the same item. When an item is repeated in another trial, the process continues to evolve from its value at the time a decision was last made for or against the item.

We fit the model to the data from Bakkour et al. (2019). The model was able to capture the relationship between choice, response time and Δve (Fig. 11A). Fig. 11B shows the degree of correlation in the evidence stream as a function of time, for the model that best fit each participant’s data. After 1 second of evidence sampling, the correlation was 0.1062 ± 0.0113 (mean ± s.e.m. across participants). This is neither negligible (which would make the model equivalent to the DDM) nor very high (which would render sequential sampling useless, since it can only average out the noise that is not shared across time).

Revaluation occurs in a DDM with temporally-correlated noise.

A drift-diffusion model with non-independent noise (ceDDM) captures the main features of revaluation. (A) The ceDDM accounts for choices (top) and response times (bottom), plotted as a function of the difference in values obtained from explicit reports (Δve). Same data as in Fig. 1C-D. Red curves are simulations of the best-fitting model. Each trial was simulated 100 times. Simulations were first averaged within trials and then averaged across trials. Error bars and bands indicate s.e.m. across trials. (B) Noise correlations as a function of time lag, obtained from the best-fitting model. Each curve corresponds to a different participant. Each curve corresponds to a different participant. (C) δ parameters derived by applying Reval to simulated data from the best fitting ceDDM model to each participant’s data. As in the data, δ > 0 for all participants. (D) Similar analysis as in Fig. 5B applied to simulations of the ceDDM. As for the data, Reval increased the range of RTs obtained after grouping trials by difficulty (by e-values on the left and r-values on the right; p-value from paired t-test). (E) Similar analysis to that of Fig. 7, using the simulated data. As observed in the data, the deviance resulting from applying Reval in the correct trial order (abscissa) is smaller than when applied in the opposite order (p-value from paired t-test).

The assumptions embodied by the ceDDM are consistent with the results of the Reval analysis. We applied the Reval algorithm to simulated data obtained from the best-fitting ceDDM. The results were in good agreement with the experimental data. The best-fitting δ values were positive for all participants and in a range similar to what we observed in the data (Fig. 11C). Reval increased the range of RTs when trials were divided by difficulty, implying that Reval led to a better classification of easy and difficult decisions (Fig. 11D). Furthermore, Reval applied to the trials in the true order explained the simulated choices better than when applied in the opposite direction (Fig. 11E). This is because the model assumes that when an item first appears, the last sample obtained for that item was the value reported in the ratings phase for that item. As more samples are obtained for a given item, the correlation with the explicit values gradually decreases. The success of ceDDM implies that the sequential dependencies we identify with Reval may be the result of a value construction process necessary to make a preferential choice.

Discussion

We identified sequential dependencies between choices in a value-based decision task. Participants performed a task in which they had to make a sequence of choices among a limited set of items. The best explanation for future choices was obtained by assuming that the subjective value of the chosen item increases and the value of the unchosen item decreases after each decision. Evidence for revaluation was obtained by analyzing the probability that participants make the same decision in pairs of trials with identical options. We also identified revaluation using an algorithm we call Reval. The same algorithm allowed us to identify revaluation in other datasets obtained with the food-choice task (Folke et al., 2016; Sepulveda et al., 2020), with results similar to those we obtained from the dataset of Bakkour et al. (2019).

The sequential effects we identified can be interpreted as a manifestation of choice-induced preference change. The usual paradigms for detecting the presence of CIPCs are based on the comparison of value ratings reported before and after a choice (for a review see Izuma and Murayama, 2013; Enisman et al., 2021). After a difficult decision, the rating of the chosen alternative often increases and that of the rejected alternative often decreases—an effect termed the “spreading of alternatives”. Many variants of the free choice paradigm have been developed to control for or eliminate the statistical artifact reported by Chen and Risen (2010). One common approach is to compare the “spreading of alternatives” observed in the free-choice paradigm (rate-choose-rate, or RCR) with a control task in which a different set of participants rate the items twice before the choice phase (RRC). Any spread observed in the RRC condition cannot be explained by the CIPC, since in the RRC condition there is no choice between the two rating phases. The CIPC is measured indirectly, as the difference in the spread of the alternatives between the RCR and the RRC. Other approaches involve asking participants to rate an item that they are led to believe they have chosen, when in fact they have not (Sharot et al., 2010; Johansson et al., 2014). Any change in ratings cannot be due to the information provided by a choice, since no real choice was made. In addition to the complications introduced by deceiving the participants (e.g., participants may suspect the deception but not mention it to the experimenter), the elimination of a real choice prevents these paradigms from being used to study the process through which subjective values undergo revision during decision formation.

In contrast, our approach to identify changes in value does not require pre- and post-choice ratings. Instead, it requires a sequence of trials in which the same items are presented multiple times (as in Luettgau et al., 2020). The revaluation effect we find cannot be explained by the artifact identified by Chen and Risen (2010). Using trials with identical items, we show that the nearer in time the trials with identical items are to each other, the more likely people are to choose the same option. Further, the revaluation algorithm explains choices better when applied in the order in which the trials were presented than when applied in the reverse order. These observations are inconsistent with the notion that item values are fixed (i.e., do not change) during the experiment, regardless of whether values are the same or different from those reported during the rating phase.

We cannot determine with certainty whether the revaluation occurs after the decision or during the deliberation process leading up to the decision. At face value, it might seem that Reval implements change after each decision (Festinger, 1957). Yet, Reval simply identifies a change in value, which may well occur during the deliberation leading to the decision, perhaps owing to a comparison of other items (on other trials) that happen to suggest a dimension of comparison that increases in importance on the current trial (Lee and Daunizeau, 2020; Lichtenstein and Slovic, 2006). More broadly, the subjective value of an option depends on the mindset of the decision maker. This internal state, which in the food-choice task includes aspects such as degree of satiety or sugar craving, can vary over time, causing the value of the items to vary as well. If changes in mindset are slow—that is, lasting longer than the duration of a decision—then the value of items will be correlated over time.

We proposed a decision model (ceDDM) in which evidence samples are correlated over time. Fitting the model to account for each participant’s choices and response times produces a revaluation of magnitude similar to what we observed experimentally. It also predicts that applying Reval in the direction in which the trials were presented explains the choices better than applying it in the opposite direction, as we observed in the data. This modeling exercise suggests that the CIPC-like effects we identified may be due to processes that occur during the deliberation leading up to a choice, rather than post-decision processes that attempt to reduce cognitive dissonance. To be clear, we interpret the ceDDM only as a proxy for a variety of more nuanced processes. If the mindset endures many individual decisions, the subjective value of an item will be correlated over time. While the ceDDM captures only a small aspect of this complex process, it has allowed us to explain the sequential dependencies we identified with Reval.

The ceDDM belongs to a class of sequential sampling models in which the drift rate varies over time. Such models have already been studied in the context of value-based decisions. For example, in the attentional drift-diffusion model (Krajbich et al., 2010), the drift rate varies depending on which item is attended, as if the value of the unattended items are discounted by a multiplicative factor. In Dynamic Field Theory (Busemeyer and Townsend, 1993), the drift rate varies depending on which attribute is attended. Recently, Lee and Pezzulo (2022) showed that a sequential sampling model in which the drift rate varies over time can explain the ‘spreading of alternatives’ (SoA) characteristic of choice-induced preference change. Lee and Pezzulo (2022) propose that the initial rating of the items may be constructed using only the most salient attributes of each item, while in a difficult decision more attributes may be considered, leading to a revaluation that informs the rating reported after the decision phase (see also Voigt et al., 2019). Consistent with our proposal, Lee and Pezzulo (2022) argue that thinking about non-prominent features during decision-making increases the likelihood that these features will be recalled when evaluating options in subsequent instances.

We observed that the degree of revaluation was higher for the chosen item than for the unchosen item. This was revealed by a variant of the Reval algorithm in which we allowed both items to have different updates. We speculate that this difference can be explained by the asymmetric distribution of attention between the chosen and unchosen items. It is known that the chosen item is looked at longer than the unchosen item (Krajbich et al., 2010). Further, CIPC is more likely for items that are remembered to have been chosen or unchosen (Salti et al., 2014). So one possibility is that the revaluation is larger for the chosen than for the unchosen item because participants spent more time looking at the chosen item and thus are more likely to remember it, leading to a larger change in value (Voigt et al., 2019).

Another possibility derives from the constructive view of preferences and the potential role of attention in decision-making. It is often assumed that value-based decisions involve gathering evidence from different alternatives, and that more evidence is gathered from alternatives that are attended to for longer (Callaway et al., 2021; Li and Ma, 2021; Krajbich et al., 2010). In the ceDDM, the correlation in value for a given item decreases with the number of evidence samples collected from the item (Fig. 11B). Therefore, the more that attention is focused on a given item, the greater the difference between the item’s value before and after the decision. Because chosen items are attended to for longer than unchosen items (e.g., Krajbich et al., 2010), the chosen item should exhibit larger revaluation than the unchosen one, which is what we observed in the data (Fig. 8).

Our research contributes to a growing body of work exploring the impact of memory on decision-making and preference formation (Biderman et al., 2020), and in particular to the CIPC. It has been suggested that the retrieval of an item’s value during decision-making renders it susceptible to modification, leading to a revaluation that influences subsequent valuations through a process that has a neural correlate in the hippocampus (Luettgau et al., 2020). The link between memorability and preference is also supported by experiments in which the presentation of an item coincides with an unrelated rapid motor response that increases subsequent preference for the item (Botvinik-Nezer et al., 2021) and by experiments demonstrating that people prefer items to which they have previously been exposed (Zajonc, 1968). As in these studies, ours also highlights the role of memory in revaluation. Due to the associative nature of memory, successive evidence samples are likely to be dependent (Rhodes and Turvey, 2007). A compelling illustration of this effect was provided by Elias Costa and colleagues (Elias Costa et al., 2009). Participants were asked to report the first word that came to mind when presented with a word generated by another participant, which was then shown to yet another participant. The resulting chain resembled Lévy flights in semantic space, characterized by mostly short transitions to nearby words and occasional large jumps. Similar dynamic processes have been used to describe eye movements during visual search (Bella-Fernández et al., 2021) and the movement of animals during reward foraging (Brown et al., 2007; Hills et al., 2015). It is intriguing to consider that a similar process may describe how decision-makers search their memory for evidence that bears on a decision.

Methods

Food choice task

A total of 30 participants completed the snack task, which consisted of a rating and a choice phase. The experimental procedures were approved by the Institutional Review Board (IRB) at Columbia University, and participants provided signed informed consent before participating in the study. The data were previously published in Bakkour et al. (2019).

Rating Phase

Participants were shown a series of snack items in a randomized order on a computer screen. They indicated their willingness to pay (WTP) by using the computer mouse to move a cursor along an analog scale ranging from $0 to $3 at the bottom of the screen. The process was self-paced, and each snack item was presented one at a time. After completing the ratings for all 60 items, participants were given the opportunity to revise their ratings. The 60 items were re-displayed in random order, with the original bids displayed below each item. Participants either chose to keep their original bid by clicking “NO” or to revise the bid by clicking “YES,” which re-displayed the analog scale for bid adjustment. We take the final WTP that is reported for each item as the corresponding explicit value (e-value).

Choice phase

From the 60 rated items, 150 unique pairs were formed, ensuring variation in Δve. Each of the 60 items was included in five different pairs. The 60 item pairs were presented twice, resulting in a total of 210 trials per participant. Item pairs were presented in random order, with one item on each side of a central fixation cross. Participants were instructed to select their preferred food item and were informed that they would receive their chosen food from a randomly selected trial to consume at the end of the experiment. The task took place in an MRI scanner. Participants indicated their choice on each trial by pressing one of two buttons on an MRI-compatible button box. They had up to 3 seconds to make their choice. Once a choice was made, the chosen item was highlighted for 500 ms. Trials were separated by an inter-trial interval (ITI) drawn from a truncated exponential distribution with a minimum ITI of 1 and a maximum ITI of 12 seconds. The resulting distribution of ITIs across trials had a true mean of 3.05 seconds and a standard deviation of 2.0 seconds.

Data analysis

Association between the e-values, choice and RT

We used the following logistic regression model to evaluate the association between the e-values and the probability of choosing the item on the right: where Ii is an indicator variable that takes the value 1 if the trial was completed by subject i and 0 otherwise. We used a t-test to evaluate the hypothesis that the corresponding regression coefficient is zero, using the standard error of the estimated regression coefficient.

Similarly, we used a linear regression model to test the influence of Δve on response times: where | ⋅ | denotes absolute value and Σve is the sum of the value of the two items presented on each trial. The last term was included to account for the potential influence of value sum on response time (Smith and Krajbich, 2019).

Predicting choices in cynosure trials

We used two logistic regression models to predict the choice in each trial using observations from the other trials. We refer to the trial under consideration as the cynosure trial (Fig. 2). One model uses the explicitly reported values: while the other model uses the choices made on other trials: Where For this model, we included an L2 regularization with λ = 0.5. Both models were fit independently for each participant. We only included trials with the first appearance of each item pair (i.e., we did not include the repeated trials) so that the choice prediction for the cynosure trial is not influenced by the choice made in the paired trial containing the same items as in the cynosure trial.

Association between r-values and choice

We tested the association between r-values and choice with a logistic regression model fit to the choices. We included separate regressors for Δvr and Δve:

Choice and response time functions

When plotting the psychometric and chronometric functions (e.g., Fig. 1C-D), we binned trials depending on the value of Δve (or Δvr). The bins are defined by the following edges: { −∞,-1.5,-0.75,-0.375,-0.1875,-0.0625, 0.0625,0.1875,0.375,0.75,1.5,∞ }. We averaged the choice or RT for the trials (grouped across participants) within each bin and plotted them aligned to the mean Δvx of each bin.

Match probability

We used logistic regression to determine if the probability of giving the same response to the pair of trials with identical stimuli depended on the number of trials in between (Fig. 3). The model is: where pmatch is the probability of choosing the same item on both occasions, Ii is an indicator variable that takes a value of 1 if the pair of trials correspond to subject i, and zero otherwise, and T1st and T2nd are the trial number of the first and second occurrences of the same pair, respectively. We used a t-test to evaluate the hypothesis that β2 = 0 (i.e., that the separation between trials with identical stimuli had no effect on pmatch.

Drift-diffusion model

We fit the choice and RT data with a drift-diffusion model. It assumes that the decision variable, x, is given by the accumulation of signal and noise, where the signal is a function of the difference in value between the items, Δv, and the noise is equal to , where dt is the time step, such that the accumulated noise after 1 second of unbounded accumulation, the variance of the accumulated noise is equal to 1. The decision variable follows the difference equation, where ηt is sampled from a normal distribution with a mean 0 and variance 1, κ is a signal-noise parameter, μ is the drift rate and μ0 is a bias coefficient that is included to account for potential asymmetries between right and left choices.

We assume that the drift rate is a (potentially nonlinear) function of Δvx. We parameterize this relationship as a power law, so that where sign is the sign operation, || indicates absolute value, and γ is a fit parameter.

The decision terminates when the accumulated evidence reaches an upper bound, signaling a rightward choice, or a lower bound, signaling a leftward choice. The bound is assumed to collapse over time. It is constant until time d, and then it collapses at rate a: Collapsing bounds are needed to explain why choices that are consistent with the value ratings are usually faster than inconsistent choices for the same Δvx.

The response time is the sum of the the decision time, given by the time taken by the diffusing particle to reach of the bounds, and a non-decision time which is assumed to be normally distributed with mean μnd and standard deviation σnd.

The model has 8 parameters: {κ, B0, a, d, γ, μ0, μnd, σnd }. The standard deviation of the non-decision times (σnd) was fixed to 0.05 s. For the fits shown in Fig. 1C-D and Fig. 5A, we fit the model to grouped data from all participants. For the analysis of variance explained (Fig. S1), we fit the model separately for each participant. The model was fit to maximize the log of the likelihood of the parameters given the single-trial choice and RT: We evaluate the likelihood by numerically solving the Fokker-Planck (FP) equation that described the dynamics of the drift-diffusion process, using the Chang-Cooper fully-implicit method (Chang and Cooper, 1970; Kiani and Shadlen, 2009; Zylberberg et al., 2016). For computational considerations, we bin the values of Δvx to multiples of $0.1. From the numerical solution of the FP equation, we obtain the distribution of decision times, which is convolved with the truncated Gaussian distribution of non-decision latencies. The truncation ensures that the non-decision times are non-negative, which could otherwise occur during the optimization process for large values of σnd. The parameter search was performed using the Bayesian Adaptive Direct Search (BADS) algorithm (Acerbi and Ma, 2017).

Revaluation algorithm

The Reval algorithm was applied to each participant independently. The values are initialized to those reported during the ratings phase. They are then revised, based on the outcome of each trial, in the order of the experiment. The value of the chosen item is increased by δ and the value of the unchosen item is decreased by the same amount. The revaluation affects future decisions in which the same item is presented.

We searched for the value of δ∗ that minimizes the deviance of the logistic regression model specified by Eq. 1. The model’s deviance is given by: where the sum is over trials and ĉi is the probability assigned to the choice on trial i obtained from the best-fitting logistic regression model.

We complemented this iterative algorithm with a second approach that estimates δ∗ using the history of choices preceding each trial. Nearly identical δ values are derived using a single logistic regression model in which the binary choice made on each trial depends on the number of times each of the two items was selected and rejected on previous trials. The model is: where, as before, Ii is an indicator variable that takes a value of 1 if the trial was completed by subject i and 0 otherwise. The key variable is Δch. It depends on the number of past trials in which the item presented on the right in the current trial was chosen and not chosen , and similarly, number of past trials in which the item presented on the left in the current trial was chosen and not chosen : The variable Δch represents the influence of past choices. The signs in Eq. 15 are such that a positive (negative) value of Δch indicates a bias toward the right (left) item. To obtain the δ∗ in units equivalent to those derived with Reval, we need to divide the regression coefficient β2,i by the sensitivity coefficient β1,i, separately for each subject i. As can be seen in Fig. S2, the values obtained with this method are almost identical to those obtained with the Reval algorithm.

Correlated-evidence DDM

The model assumes that at each moment during the decision-making process, the decision-maker can only access a noisy sample of the value of each item. These samples are normally distributed, with parameters such that their unbounded accumulation over one second is also normally distributed with a mean equal to κve, where ve is the explicit value reported during the Ratings phase and κ is a measure of signal-to-noise, and a standard deviation equal to 1.

Crucially, for each item, the noise in successive samples is correlated. To generate the correlated samples, we sample from a Markov chain using the Metropolis-Hastings algorithm (Chib and Greenberg, 1995). The target distribution is the normally distributed value function described in the previous paragraph. The proposal density is also normally distributed. Its width determines the degree of correlation between consecutive samples. Typically, the correlation between successive samples is considered a limitation of the Metropolis-Hastings algorithm. Here, however, it allows us to generate correlated samples from a target distribution. The standard deviation of the proposal density is . Higher values of τ result in a narrower proposal density, hence more strongly correlated samples. We sample from the same Markov chain across different trials in which the same item is presented, so that the last sample obtained about an item in a given trial is the initial state of the Markov chain the next time the item is presented.

At each moment (dt = 40ms), we sample one value for the left item and another for the right item, compute their difference (right minus left), and accumulate this difference until it crosses a threshold at +B0, signaling a rightward choice, or at −B0, signaling a leftward choice. The decision time is added to the non-decision time, μnd, to obtain the response time.

We fit the model to the data as follows. For each item, we simulate many Markov chains. In each trial, i, we take samples from each chain until the accumulation of these samples reaches one of the two decision thresholds. Then we calculate the likelihood (L) of obtaining the choice and the RT displayed by the participant on that trial as: where N = 1, 000 is the number of Markov chains, 1 is an indicator function that takes the value 1 if the choice made on chain j is the same as the choice made by the participant on trial i and 0 otherwise, N (x|y, z) is the normal probability density function with mean y and standard deviation z evaluated at x, and σnd is a parameter fit to the data.

When an item is presented again in a future trial, the initial state of each Markov chain depends on the state it was in the last time the item was presented. The initial state of each chain is obtained by sampling 1,000 values (one per chain) from the distribution given by the final state of each chain. The sampling is weighted by the value of Lj of each chain (Eq. 16), so that chains that better explained the choice and RT in the last trial are more likely to be sampled from in future trials.

The model has 5 parameters per participant: κ, B0, τ, μnd, σnd, which were fit to maximize the sum, across trials, of the log of L using BADS (Acerbi and Ma, 2017).

fMRI analysis

Acquisition

Imaging data were acquired on a 3T GE MR750 MRI scanner with a 32-channel head coil. Functional data were acquired using a T2∗-weighted echo planar imaging sequence (repetition time (TR) = 2 s, echo time (TE) = 22 ms, flip angle (FA) = 70◦, field of view (FOV) = 192 mm, acquisition matrix of 96 x 96). Forty oblique axial slices were acquired with a 2 mm in-plane resolution positioned along the anterior commissure-posterior commissure line and spaced 3 mm to achieve full brain coverage. Slices were acquired in an interleaved fashion. We acquired three runs of the food choice task, each composed of 70 trials. Each of the food choice task functional runs consisted of 212 volumes and lasted 7 minutes. In addition to functional data, a single three-dimensional high-resolution (1 mm isotropic) T1-weighted full-brain image was acquired using a BRAVO pulse sequence for brain masking and image registration.

Preprocessing

Raw imaging data in DICOM format were converted to NIFTI format and preprocessed through a standard preprocessing pipeline using the Oxford Centre for Functional Magnetic Resonance Imaging of the Brain (FMRIB) Software Library (FSL) package version 5 (Smith et al., 2004). Functional image time series were first aligned via Motion Correction using FMRIB’s Linear Image Registration Tool (MCFLIRT) to obtain six motion parameters that correspond to the x- y-, and z-axis translations and rotations of the brain over time. Then, the skull was removed from the T2∗ images using the Brain Extraction tool (BET) and from the high-resolution T1 images using Freesurfer (Fischl et al., 1999; Ségonne et al., 2004). Spatial smoothing was performed using a Gaussian kernel with a full-width half maximum (FWHM) of 5 mm. Data and design matrix were high-pass filtered using a Gaussian-weighted least-squares straight line fit with a cutoff period of 100 s. Grand-mean intensity normalization of each run’s entire four-dimensional data set by a single multiplicative factor was performed. The functional volumes for each participant and run were registered to the high resolution T1-weighted structural volume using a non-linear boundary-based registration method implemented in FSL5 (Greve and Fischl, 2009). The T1-weighted image was then registered to the MNI152 2 mm template using FMRIB’s Linear Image Registration Tool (FLIRT, 12 degrees of freedom). These two registration steps were concatenated to obtain a functional-to-standard space registration matrix.

Analysis

We conducted a GLM analysis to look at BOLD activity related to r-values and e-values. This model included eight regressors: (i) onsets for all valid trials, modeled with a duration equal to the average RT across all valid choices and participants; (ii) same onsets and duration as (i) but modulated by |Δve | de-meaned across these trials within each run for each participant; (iii) same onsets and duration as (i) but modulated by |Δvr | demeaned across these trials within each run for each participant; (iv) same onsets and duration as (i) but modulated by RT demeaned across these trials within each run for each participant; (v) same onsets and duration as (i) but modulated by the e-value of the chosen item demeaned across trials within each run for each participant; (vi) same onsets and duration as (i) but modulated by the r-value of the chosen item demeaned across these trials within each run for each participant; (vii) to account for any differences in right/left choices between trial types we added a regressor with the same onsets and durations as (i), while the modulator was an indicator for right/left response; (viii) onsets for missed trials. The map in Fig. 9 was generated using this model.

The model includes the six x, y, z translation and rotation motion parameters obtained from MCFLIRT, framewise displacement (FD) and RMS intensity difference from one volume to the next (Power et al., 2012) as confounding regressors. We also modelled volumes with FD and DVARS exceeding a threshold of 0.5 by adding a single time point regressor for each ‘to be scrubbed’ volume (Siegel et al., 2014). All regressors were entered at the first level of analysis, and all (except the added confounding regressors) were convolved with a canonical double-gamma hemodynamic response function. The time derivative of each regressor (except the added confounding regressors) was included in the model. Models were estimated separately for each participant and run.

GLMs were estimated using FSL’s FMRI Expert Analysis Tool (FEAT). The first-level time-series GLM analysis was performed for each run per participant using FSL’s FILM. The first-level contrast images were then combined across runs per participant using fixed effects. The group-level analysis was performed using FMRIB’s Local Analysis of Mixed Effects (FLAME1) (Beckmann et al., 2003). Group-level maps were corrected to control the family-wise error rate using cluster-based Gaussian random field correction for multiple comparisons, with an uncorrected cluster-forming threshold of z=2.3 and corrected extent threshold of p < 0.05.

Author contributions

The data were collected and published by Bakkour et al. (2019). AZ conceived and designed the present study, performed the analyses, implemented the models, and wrote a draft of the manuscript. AB conducted the fMRI analysis. All authors helped to revise the final manuscript. DS and MNS provided intellectual support throughout the study.

eLife Assessment

The authors address key assumptions underlying current models of the formation of value-based decisions. They provide solid evidence that the subjective values human participants assign to choice options change across sequences of multiple decisions and establish valuable methods to detect these changes in frequently used behavioral task designs. That said, the description of the fMRI results requires further elaboration in order to support the claim that the authors' algorithm reveals neural valuation processes better than the current standard approach.

Reviewer #1 (Public Review):

Summary:

There is a long-standing idea that choices influence evaluation: options we choose are re-evaluated to be better than they were before the choice. There has been some debate about this finding, and the authors developed several novel methods for detecting these re-evaluations in task designs where options are repeatedly presented against several alternatives. Using these novel methods the authors clearly demonstrate this re-evaluation phenomenon in several existing datasets.

Strengths:

The paper is well-written and the figures are clear. The authors provided evidence for the behaviour effect using several techniques and generated surrogate data (where the ground truth is known) to demonstrate the robustness of their methods.

Weaknesses:

The description of the results of the fMRI analysis in the text is not complete: weakening the claim that their re-evaluation algorithm better reveals neural valuation processes.

Reviewer #2 (Public Review):

Summary:

Zylberberg and colleagues show that food choice outcomes and BOLD signal in the vmPFC are better explained by algorithms that update subjective values during the sequence of choices compared to algorithms based on static values acquired before the decision phase. This study presents a valuable means of reducing the apparent stochasticity of choices in common laboratory experiment designs. The evidence supporting the claims of the authors is solid, although currently limited to choices between food items because no other goods were examined. The work will be of interest to researchers examining decision-making across various social and biological sciences.

Strengths:

The paper analyses multiple food choice datasets to check the robustness of its findings in that domain.

The paper presents simulations and robustness checks to back up its core claims.

Weaknesses:

To avoid potential misunderstandings of their work, I think it would be useful for the authors to clarify their statements and implications regarding the utility of item ratings/bids (e-values) in explaining choice behavior. Currently, the paper emphasizes that e-values have limited power to predict choices without explicitly stating the likely reason for this limitation given its own results or pointing out that this limitation is not unique to e-values and would apply to choice outcomes or any other preference elicitation measure too. The core of the paper rests on the argument that the subjective values of the food items are not stored as a relatively constant value, but instead are constructed at the time of choice based on the individual's current state. That is, a food's subjective value is a dynamic creation, and any measure of subjective value will become less accurate with time or new inputs (see Figure 3 regarding choice outcomes, for example). The e-values will change with time, choice deliberation, or other experiences to reflect the change in subjective value. Indeed, most previous studies of choice-induced preference change, including those cited in this manuscript, use multiple elicitations of e-values to detect these changes. It is important to clearly state that this paper provides no data on whether e-values are more or less limited than any other measure of eliciting subjective value. Rather, the paper shows that a static estimate of a food's subjective value at a single point in time has limited power to predict future choices. Thus, a more accurate label for the e-values would be static values because stationarity is the key assumption rather than the means by which the values are elicited or inferred.

There is a puzzling discrepancy between the fits of a DDM using e-values in Figure 1 versus Figure 5. In Figure 1, the DDM using e-values provides a rather good fit to the empirical data, while in Figure 5 its match to the same empirical data appears to be substantially worse. I suspect that this is because the value difference on the x-axis in Figure 1 is based on the e-values, while in Figure 5 it is based on the r-values from the Reval algorithm. However, the computation of the value difference measure on the two x-axes is not explicitly described in the figures or methods section and these details should be added to the manuscript. If my guess is correct, then I think it is misleading to plot the DDM fit to e-values against choice and RT curves derived from r-values. Comparing Figures 1 and 5, it seems that changing the axes creates an artificial impression that the DDM using e-values is much worse than the one fit using r-values.

Relatedly, do model comparison metrics favor a DDM using r-values over one using e-values in any of the datasets tested? Such tests, which use the full distribution of response times without dividing the continuum of decision difficulty into arbitrary hard and easy bins, would be more convincing than the tests of RT differences between the categorical divisions of hard versus easy.

Revaluation and reduction in the imprecision of subjective value representations during (or after) a choice are not mutually exclusive. The fact that applying Reval in the forward trial order leads to lower deviance than applying it in the backwards order (Figure 7) suggests that revaluation does occur. It doesn't tell us if there is also a reduction in imprecision. A comparison of backwards Reval versus no Reval would indicate whether there is a reduction in imprecision in addition to revaluation. Model comparison metrics and plots of the deviance from the logistic regression fit using e-values against backward and forward Reval models would be useful to show the relative improvement for both forms of Reval.

Did the analyses of BOLD activity shown in Figure 9 orthogonalize between the various e-value- and r-value-based regressors? I assume they were not because the idea was to let the two types of regressors compete for variance, but orthogonalization is common in fMRI analyses so it would be good to clarify that this was not used in this case. Assuming no orthogonalization, the unique variance for the r-value of the chosen option in a model that also includes the e-value of the chosen option is the delta term that distinguishes the r and e-values. The delta term is a scaled count of how often the food item was chosen and rejected in previous trials. It would be useful to know if the vmPFC BOLD activity correlates directly with this count or the entire r-value (e-value + delta). That is easily tested using two additional models that include only the r-value or only the delta term for each trial.

Please confirm that the correlation coefficients shown in Figure 11 B are autocorrelations in the MCMC chains at various lags. If this interpretation is incorrect, please give more detail on how these coefficients were computed and what they represent.

The paper presents the ceDDM as a proof-of-principle type model that can reproduce certain features of the empirical data. There are other plausible modifications to bounded evidence accumulation (BEA) models that may also reproduce these features as well or better than the ceDDM. For example, a DDM in which the starting point bias is a function of how often the two items were chosen or rejected in previous trials. My point is not that I think other BEA models would be better than the ceDDM, but rather that we don't know because the tests have not been run. Naturally, no paper can test all potential models and I am not suggesting that this paper should compare the ceDDM to other BEA processes. However, it should clearly state what we can and cannot conclude from the results it presents.

This work has important practical implications for many studies in the decision sciences that seek to understand how various factors influence choice outcomes. By better accounting for the context-specific nature of value construction, studies can gain more precise estimates of the effects of treatments of interest on decision processes. That said, there are limitations to the generalizability of these findings that should be noted.

These limitations stem from the fact that the paper only analyzes choices between food items and the outcomes of the choices are not realized until the end of the study (i.e., participants do not eat the chosen item before making the next choice). This creates at least two important limitations. First, preferences over food items may be particularly sensitive to mindsets/bodily states. We don't yet know how large the choice deltas may be for other types of goods whose value is less sensitive to satiety and other dynamic bodily states. Second, the somewhat artificial situation of making numerous choices between different pairs of items without receiving or consuming anything may eliminate potential decreases in the preference for the chosen item that would occur in the wild outside the lab setting. It seems quite probable that in many real-world decisions, the value of a chosen good is reduced in future choices because the individual does not need or want multiples of that item. Naturally, this depends on the durability of the good and the time between choices. A decrease in the value of chosen goods is still an example of dynamic value construction, but I don't see how such a decrease could be produced by the ceDDM.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

There is a long-standing idea that choices influence evaluation: options we choose are re-evaluated to be better than they were before the choice. There has been some debate about this finding, and the authors developed several novel methods for detecting these re-evaluations in task designs where options are repeatedly presented against several alternatives. Using these novel methods the authors clearly demonstrate this re-evaluation phenomenon in several existing datasets.

Strengths:

The paper is well-written and the figures are clear. The authors provided evidence for the behaviour effect using several techniques and generated surrogate data (where the ground truth is known) to demonstrate the robustness of their methods.

Weaknesses:

The description of the results of the fMRI analysis in the text is not complete: weakening the claim that their re-evaluation algorithm better reveals neural valuation processes.

We appreciate the reviewer’s comment regarding the incomplete account of the fMRI results. In response, we implemented Reviewer #2's suggestion to run additional GLM models for a clearer interpretation of our findings. We also took this opportunity to apply updated preprocessing to the fMRI data and revise the GLM models, making them both simpler and more comprehensive. The results section is thus substantially revised, now including a new main figure and several supplemental figures that more clearly present our fMRI findings. Additionally, we have uploaded the statistical maps to NeuroVault, allowing readers to explore the full maps interactively rather than relying solely on the static images in the paper. The new analyses strengthen our original conclusion: dynamic values (previously referred to as revalued values, following the reviewer’s suggestion) better explain BOLD activity in the ventromedial prefrontal cortex, a region consistently associated with valuation, than static values (values reported prior to the choice phase in the auction procedure).

Reviewer #2 (Public Review):

Summary:

Zylberberg and colleagues show that food choice outcomes and BOLD signal in the vmPFC are better explained by algorithms that update subjective values during the sequence of choices compared to algorithms based on static values acquired before the decision phase. This study presents a valuable means of reducing the apparent stochasticity of choices in common laboratory experiment designs. The evidence supporting the claims of the authors is solid, although currently limited to choices between food items because no other goods were examined. The work will be of interest to researchers examining decision-making across various social and biological sciences.

Strengths:

The paper analyses multiple food choice datasets to check the robustness of its findings in that domain.

The paper presents simulations and robustness checks to back up its core claims.

Weaknesses:

To avoid potential misunderstandings of their work, I think it would be useful for the authors to clarify their statements and implications regarding the utility of item ratings/bids (e-values) in explaining choice behavior. Currently, the paper emphasizes that e-values have limited power to predict choices without explicitly stating the likely reason for this limitation given its own results or pointing out that this limitation is not unique to e-values and would apply to choice outcomes or any other preference elicitation measure too. The core of the paper rests on the argument that the subjective values of the food items are not stored as a relatively constant value, but instead are constructed at the time of choice based on the individual's current state. That is, a food's subjective value is a dynamic creation, and any measure of subjective value will become less accurate with time or new inputs (see Figure 3 regarding choice outcomes, for example). The e-values will change with time, choice deliberation, or other experiences to reflect the change in subjective value. Indeed, most previous studies of choice-induced preference change, including those cited in this manuscript, use multiple elicitations of e-values to detect these changes. It is important to clearly state that this paper provides no data on whether e-values are more or less limited than any other measure of eliciting subjective value. Rather, the paper shows that a static estimate of a food's subjective value at a single point in time has limited power to predict future choices. Thus, a more accurate label for the e-values would be static values because stationarity is the key assumption rather than the means by which the values are elicited or inferred.

Thank you for this helpful comment. We changed the terminology following the reviewer’s suggestion. The “explicit” values (e-values or ve) are now called “static” values (s-values or vs). Accordingly, we also changed the “Reval” values (r-values or vr) to “dynamic” values (d-values or vd).

We also address the reviewer's more general point about the utility of item ratings/bids (s-values) and whether our results are likely to hold with other ways of eliciting subjective values. We added a new sub-section in Discussion addressing this and other limitations of our study. To address the reviewer’s point, we write:

“One limitation of our study is that we only examined tasks in which static values were elicited from explicit reports of the value of food items. It remains to be determined if other ways of eliciting subjective values (e.g., Jensen and Miller, 2010) would lead to similar results. We think so, as the analysis of trials with identical item pairs (Fig. 3) and the difference between forward and backward Reval (Fig. 7) are inconsistent with the notion that values are static, regardless of their precise value. It also remains to be determined if our results will generalize to non-food items whose value is less sensitive to satiety and other dynamic bodily states. Perceptual decisions also exhibit sequential dependencies, and it remains to be explored whether these can be explained as a process of value construction, similar to what we propose here for the food-choice task (Gupta et al., 2024; Cho et al., 2002; Zylberberg et al., 2018; Abrahamyan et al., 2016).”

There is a puzzling discrepancy between the fits of a DDM using e-values in Figure 1 versus Figure 5. In Figure 1, the DDM using e-values provides a rather good fit to the empirical data, while in Figure 5 its match to the same empirical data appears to be substantially worse. I suspect that this is because the value difference on the x-axis in Figure 1 is based on the e-values, while in Figure 5 it is based on the r-values from the Reval algorithm. However, the computation of the value difference measure on the two x-axes is not explicitly described in the figures or methods section and these details should be added to the manuscript. If my guess is correct, then I think it is misleading to plot the DDM fit to e-values against choice and RT curves derived from r-values. Comparing Figures 1 and 5, it seems that changing the axes creates an artificial impression that the DDM using e-values is much worse than the one fit using r-values.

We agree with the reviewer that this way of presenting the DDM fits could be misleading. In the previous version of the manuscript, we included the two fits in the same figure panel to make it clear that the sensitivity (slope) of the choice function is greater when we fit the data using the r-values (now d-values) than when we fit them using the e-values (now s-values). In the revised version of Figure 5, we include the data points already shown in Figure 1, so that each DDM fit is shown with their corresponding data points. Thus we avoid giving the false impression that the DDM model fit using the s-values is much worse than the one fit using the d-values. This said, the fit is indeed worse, as we now show with the formal model comparison suggested by the reviewer (next comment).

Relatedly, do model comparison metrics favor a DDM using r-values over one using e-values in any of the datasets tested? Such tests, which use the full distribution of response times without dividing the continuum of decision difficulty into arbitrary hard and easy bins, would be more convincing than the tests of RT differences between the categorical divisions of hard versus easy.

We now include the model comparison suggested by the reviewer. The comparison shows that the DDM model using dynamic values explains the choice and response time data better than one using static values. One potential caveat of this comparison, which explains why we did not include it in the original version of the manuscript, is that the d-values are obtained from a fit to the choice data, which could bias the subsequent DDM comparison. We control for this in three ways: (1) by calculating the difference in Bayesian Information Criterion (BIC) between the models, penalizing the DDM model that uses the d-values for the additional parameter (δ); (2) by comparing the difference in BIC against simulations of a model in which the choice and RT data were obtained assuming static values; this analysis shows that if values were static, the DDM using static values would be favored in the comparison despite having one fewer parameter; (3) ignoring the DDM fit to the choices in the model comparison, and just comparing how well the two models explain the RTs; this comparison is unbiased because the δ values are fit only to the choice data, not the RTs. These analyses are now included in Figure 5 and Figure 5–Figure supplement 2.

Revaluation and reduction in the imprecision of subjective value representations during (or after) a choice are not mutually exclusive. The fact that applying Reval in the forward trial order leads to lower deviance than applying it in the backwards order (Figure 7) suggests that revaluation does occur. It doesn't tell us if there is also a reduction in imprecision. A comparison of backwards Reval versus no Reval would indicate whether there is a reduction in imprecision in addition to revaluation. Model comparison metrics and plots of the deviance from the logistic regression fit using e-values against backward and forward Reval models would be useful to show the relative improvement for both forms of Reval.

We agree with the reviewer that the occurrence of revaluation does not preclude other factors from affecting valuation. Following the reviewer’s suggestion we added a panel to Figure 6 (new panel B), in which we show the change in the deviance from the logistic regression fits between Reval (forward direction) and no-Reval. The figure clearly shows that the difference in deviance for the data is much larger than that obtained from simulations of choice data generated from the logistic fits to the static values (shown in red).

Interestingly, we also observe that the deviance obtained after applying Reval in the backward direction is lower than that obtained using the s-values. We added a panel to figure 7 showing this (Fig. 7B). This observation, however, does not imply that there are factors affecting valuation besides revaluation (e.g.,”reduction in imprecision”). Indeed, as we now show in a new panel in Figure 11 (panel F), the same effect (lower deviance for backward Reval than no-Reval) is observed in simulations of the ceDDM.

Besides the new figure panels (Fig. 6B, 7B, 11F), we mention in Discussion (new subsection, “Limitations...”, paragraph #2) the possibility that there are other non-dynamic contributions to the reduction in deviance for Backward Reval compared to no-Reval:

“Another limitation of our study is that, in one of the datasets we analyzed (Sepulveda et al. 2020), applying Reval in the forward direction was no better than applying it in the backward direction (Fig. 10). We speculate that this failure is related to idiosyncrasies of the experimental design, in particular, the use of alternating blocks of trials with different instructions (select preferred vs. select non-preferred). More importantly, Reval applied in the backward direction led to a significant reduction in deviance relative to that obtained using the static values. This reduction was also observed in the ceDDM, suggesting that the effect may be explained by the changes in valuation during deliberation. However, we cannot discard a contribution from other, non-dynamic changes in valuation between the rating and choice phase including contextual effects (Lichtenstein and Slovic, 2006), stochastic variability in explicit value reporting (Polania et al., 2019), and the limited range of numerical scales used to report value.”

Did the analyses of BOLD activity shown in Figure 9 orthogonalize between the various e-valueand r-value-based regressors? I assume they were not because the idea was to let the two types of regressors compete for variance, but orthogonalization is common in fMRI analyses so it would be good to clarify that this was not used in this case. Assuming no orthogonalization, the unique variance for the r-value of the chosen option in a model that also includes the e-value of the chosen option is the delta term that distinguishes the r and e-values. The delta term is a scaled count of how often the food item was chosen and rejected in previous trials. It would be useful to know if the vmPFC BOLD activity correlates directly with this count or the entire r-value (e-value + delta). That is easily tested using two additional models that include only the r-value or only the delta term for each trial.

We did not orthogonalize the static value and dynamic value regressors. We have included this detail in the revised methods. We thank the reviewer for the suggestion to run additional models to improve our ability to interpret our findings. We have substantially revised all fMRI-related sections of the paper. We took this opportunity to apply standardized and reproducible preprocessing steps implemented in fmriprep, present whole-brain corrected maps on a reconstructed surface of a template brain, and include links to the full statistical maps for the reader to navigate the full map, rather than rely on the static image in the figures. We implemented four models in total: model 1 includes both static value (Vs) obtained during the auction procedure prior to the choice phase and dynamic value (Vd) output by the revaluation algorithm (similar to the model presented in the first submission); model 2 includes only delta = Vd - Vs; model 3 includes only Vs; model 4 includes only Vd. All models included the same confound and nuisance regressors. We found that Vd was positively related to BOLD in vmPFC when accounting for Vs, correcting for familywise error rate at the whole brain level. Interestingly, the relationship between delta and vmPFC BOLD did not survive whole-brain correction and the effect size of the relationship between Vd and vmPFC bold in model 4 was larger than the effect size of the relationship between Vs and vmPFC bold in model 3 and survived correction at the whole brain level encompassing more of the vmPFC. Together, these findings bolster our claim that Vd better accounts for BOLD variability in vmPFC, a brain region reliably linked to valuation.

Please confirm that the correlation coefficients shown in Figure 11 B are autocorrelations in the MCMC chains at various lags. If this interpretation is incorrect, please give more detail on how these coefficients were computed and what they represent.

We added a paragraph in Methods explaining how we compute the correlations in Figure 11B (last paragraph of the sub-section “Correlated-evidence DDM” in Methods):

“The correlations in Fig. 11B were generated using the best-fitting parameters for each participant to simulate 100,000 Markov chains. We generate Markov chain samples independently for the left and right items over a 1-second period. To illustrate noise correlations, the simulations assume that the static value of both the left and right items is zero. We then and for each of the Markov chains (𝑥). Pearson's𝑥 correlation is computed between these 𝑡 calculate the difference in dynamic value ( ) between the left and right items at each time (𝑡) differences at time zero, 𝑥𝑖(𝑡 = 0), and at time 𝑥𝑖(𝑡 = τ), for different time lags τ. Correlations were calculated independently for each participant. Each trace in Fig. 11B represents a different participant.”

The paper presents the ceDDM as a proof-of-principle type model that can reproduce certain features of the empirical data. There are other plausible modifications to bounded evidence accumulation (BEA) models that may also reproduce these features as well or better than the ceDDM. For example, a DDM in which the starting point bias is a function of how often the two items were chosen or rejected in previous trials. My point is not that I think other BEA models would be better than the ceDDM, but rather that we don't know because the tests have not been run. Naturally, no paper can test all potential models and I am not suggesting that this paper should compare the ceDDM to other BEA processes. However, it should clearly state what we can and cannot conclude from the results it presents.

Indeed, the ceDDM should be interpreted as a proof-of-principle model, which shows that drifting values can explain many of our results. It is definitely wrong in the details, and we are open to the possibility that a different way of introducing sequential dependencies between decisions may lead to a better match to the experimental data. We now mention this in a new subsection of Discussion, “Limitations...” paragraph #3:

“Finally, we emphasize that the ceDDM should be interpreted as a proof-of-principle model used to illustrate how stochastic fluctuations in item desirability can explain many of our results. We chose to model value changes following an MCMC process. However, other stochastic processes or other ways of introducing sequential dependencies (e.g., variability in the starting point of evidence accumulation) may also explain the behavioral observations. Furthermore, there likely are other ways to induce changes in the value of items other than through past decisions. For example, attentional manipulations or other experiences (e.g., actual food consumption) may change one's preference for an item. The current version of the ceDDM does not allow for these influences on value, but we see no fundamental limitation to incorporating them in future instantiations of the model.”

This work has important practical implications for many studies in the decision sciences that seek to understand how various factors influence choice outcomes. By better accounting for the context-specific nature of value construction, studies can gain more precise estimates of the effects of treatments of interest on decision processes.

Thank you!

That said, there are limitations to the generalizability of these findings that should be noted.

These limitations stem from the fact that the paper only analyzes choices between food items and the outcomes of the choices are not realized until the end of the study (i.e., participants do not eat the chosen item before making the next choice). This creates at least two important limitations. First, preferences over food items may be particularly sensitive to mindsets/bodily states. We don't yet know how large the choice deltas may be for other types of goods whose value is less sensitive to satiety and other dynamic bodily states. Second, the somewhat artificial situation of making numerous choices between different pairs of items without receiving or consuming anything may eliminate potential decreases in the preference for the chosen item that would occur in the wild outside the lab setting. It seems quite probable that in many real-world decisions, the value of a chosen good is reduced in future choices because the individual does not need or want multiples of that item. Naturally, this depends on the durability of the good and the time between choices. A decrease in the value of chosen goods is still an example of dynamic value construction, but I don't see how such a decrease could be produced by the ceDDM.

These are all great points. The question of how generalizable our results are to other domains is wide open. We do have preliminary evidence suggesting that in a perceptual decision-making task with two relevant dimensions (motion and color; Kang, Loffler et al. eLife 2021), the dimension that was most informative to resolve preference in the past is prioritized in future decisions. We believe that a similar process underlies the apparent change in value in value-based decisions. We decided not to include this experiment in the manuscript, as it would make the paper much longer and the experimental designs are very different. Exploring the question of generality is a matter for future studies.

We also agree that food consumption is likely to change the value of the items. For example, after eating something salty we are likely to want something to drink. We mention in the revised manuscript that time, choice deliberation, attentional allocation and other experiences (including food consumption) are likely to change the value of the alternatives and thus affect future choices and valuations.

The ceDDM captures only sequential dependencies that can be attributed to values that undergo diffusion-type changes during deliberation. While the ceDDM captures many of the experimental observations, the value of an item may change for reasons not captured by the ceDDM. For example, food consumption is likely to change the value of items (e.g., wanting something to drink after eating something salty). The reviewer is correct that the current version of ceDDM could not account for these changes in value. However, we see no fundamental limitation to extending the ceDDM to account for them.

We discuss these issues in a new subsection in Discussion (“Limitations...” paragraph #3).

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Summary

The authors address assumptions of bounded accumulation of evidence for value-based decision-making. They provide convincing evidence that subjects drift in their subjective preferences across time and demonstrate valuable methods to detect these drifts in certain task designs.

My specific comments are intended to assist the authors with making the paper as clear as possible. My only major concern is with the reporting of the fMRI results.

Thank you, please see our responses above for a description of the changes we made to the fMRI analyses.

Specific comments

- In the intro, I would ask the authors to consider the idea that things like slow drift in vigilance/motivation or faster drifts in spatial attention could also generate serial dependencies in perceptual tasks. I think the argument that these effects are larger in value-based tasks is reasonable, but the authors go a bit too far (in my opinion) arguing that similar effects do not exist *at all* in perceptual decision-making.

We added a sentence in the Discussion (new section on Limitations, paragraph #1) mentioning some of the literature on sequential dependencies in perceptual tasks and asking whether there might be a common explanation for such dependencies for perceptual and value-based decisions. We tried including this in the Introduction, but we thought it disrupted the flow too much.

- Figure 1: would it not be more clear to swap the order of panels A and B? Since B comes first in the task?

We agree, we swapped the order of panels A and B.

- Figure 2: the label 'simulations' might be better as 'e-value simulations'

Yes, we changed the label ‘simulations’ to ‘simulations with s-values’ (we changed the term explicit value to static value, following a suggestion by Reviewer #2).

- For the results related to Figure 2, some citations related to gaps between "stated versus revealed preferences" seem appropriate.

We added a few relevant citations where we explain the results related to Figure 2.

- Figure 3: in addition to a decrease in match preferences over the session, it would be nice to look at other features of the task which might have varied over the session. e.g. were earlier trials more likely to be predicted by e-value?

We do see a trend in this direction, but the effect is not significant. The following figure shows the consistency of the choices with the stated values, as a function of the |∆value|, for the first half (blue) and the second half (red) of the trials. The x-axis discretizes the absolute value of the difference in static value between the left and right items, binned in 17 bins of approximately equal number of trials.

The slope is shallower for the second half, but a logistic regression model revealed that the difference is not significant:

,

where Ilate is an indicator variable that takes a value of 1 for the second half of the trials and zero otherwise.

As expected from the figure β2 was negative (-0.15) but the effect was not significant (p-value =0.32, likelihood ratio test).

We feel we do not have much to say about this result, which may be due to lack of statistical power, so we would rather not include this analysis in the revised manuscript.

It is worth noting that if we repeat the analysis using the dynamic values obtained from Reval instead of the static values, the consistency is overall much greater and little difference is observed between the first and second halves of the experiment:

- The e-value DDM fit in Figure 1C/D goes through the points pretty well, but the e-value fits in 5A do not because of a mismatch with the axis. The x-axis needs to say whether the value difference is the e-value or the r-value. Also, it seems only fair to plot the DDM for the r-value on a plot with the x-axis being the e-value.

Thank you for this comment, we have now changed Figure 5A, such that both sets of data points are shown (data grouped by both e-values and by r-values). We agree that the previous version made it seem as if the fits were worse for the DDM fit to the e-values. The fits are indeed worse, as revealed by a new DDM model comparison (Figure 5–Figure supplement 2), but the effect is more subtle than the previous version of the figure implied.

- How is Figure 5B "model free" empirical support? The fact that the r-value model gives better separation of the RTs on easy and hard trials doesn't seem "model-free" and also it isn't clear how this directly relates to being a better model. It seems that just showing a box-plot of the R2 for the RT of the two models would be better?

We agree that “model free” may not be the best expression, since the r-values (now d-values) are derived from a model (Reval). Our intention was to make clear that because Reval only depends on the choices, the relationship between RT and ∆vdynamic is a prediction. We no longer use the term, model free, in the caption. We tried to clarify the point in Results, where we explain this figure panel. We have also included a new model comparison (Figure 5–Figure supplement 2), showing that the DDM model fit to the d-values explains choice and RT better than one fit to the s-values.

This said, we do consider the separation in RTs between easy and hard trials to be a valid metric to compare the accuracy of the static and dynamic values. The key assumption is that there is a monotonically decreasing relationship between value difference, ∆v, and response time. The monotonic relationship does not need to hold for individual trials (due to the noisiness of the RTs) but should hold if one were to average a large enough number of trials for each value of ∆v.

Under this assumption, the more truthful a value representation is (i.e., the closer the value we infer is to the true subjective value of the item on a given trial, assuming one exists), the greater the difference in RTs between trials judged to be difficult and those considered easy. To illustrate this with an extreme case, if an experimenter’s valuation of the items is very inaccurate (e.g., done randomly), then on average there will be no difference between easy and difficult RTs as determined by this scoring.

- Line 189: Are the stats associated with Eq 7, was the model fit subject by subject? Combining subjects? A mixed-effects model? Why not show a scatter plot of the coefficients of Δvₑ and Δvᵣ (1 point/subject).

The model was not fit separately for each subject. Instead, we concatenated trials from all subjects, allowing each subject to have a different bias term (β0,i ).

We have now replaced it with the analysis suggested by the reviewer. We fit the logistic regression model independently for each participant. The scatter plot suggested by the reviewer is shown in Figure 5–Figure supplement 1. Error bars indicate the s.e. of the regression coefficients:

It can be seen that the result is consistent with what we reported before: βd is significantly positive for all participants, while βs is not.

- I think Figure S1 should be a main figure.

Thank you for this suggestion, we have now included the former Figure S1 as an additional panel in Figure 5.

- Fig 9 figure and text (line 259) don't exactly match. In the text it says that the BOLD correlated with vᵣ and not vₑ, but the caption says there were correlations with vᵣ after controlling for vₑ. Is there really nothing in the brain that correlated with vₑ? This seems hard to believe given how correlated the two estimates are. In the methods, 8 regressors are described. A more detailed description of the results is needed.

Thank you for pointing out the inconsistency in our portrayal of the results in the main text and in the figure caption. We have substantially revised all fMRI methods, re-ran fMRI data preprocessing and implemented new, simpler, and more comprehensive GLM models following Reviewer #2's suggestion. Consequently, we have replaced Figure 9, added Figure 9 — Figure Supplement 1, and uploaded all maps to NeuroVault. These new models and maps allow for a clearer interpretation of our findings. More details about the fMRI analyses in the methods and results are included in the revision. We took care to use similar language in the main text and in the figure captions to convey the results and interpretation. The new analyses strengthen our original conclusion: dynamic values better explain BOLD activity in the ventromedial prefrontal cortex, a region consistently associated with valuation, than static values.

- It's great that the authors reanalyzed existing datasets (fig 10). I think the ΔRT plots are the least clear way to show that _reval_ is better. Why not a figure like Figure 6a and Figure 7 for the existing datasets?

We agree with the reviewer. We have replaced Fig. 10 with a more detailed version. For each dataset, we show the ΔRT plots, but we also show figures equivalent to Fig. 6a, Fig. 7a, and the new Fig. 6b (Deviance with and without Reval).

Reviewer #2 (Recommendations For The Authors):

I assume that the data and analysis code will be made publicly and openly available once the version of record is established.

Yes, the data and analysis code is now available at: https://github.com/arielzylberberg/Reval_eLife_2024

We added a Data Availability statement to the manuscript.

Version 2

Abstract

Abstract

Many decisions are expressed as a preference for one item over another. When these items are familiar, it is often assumed that the decision maker assigns a value to each of the items and chooses the item with the highest value. These values may be imperfectly recalled, but are assumed to be stable over the course of an interview or psychological experiment. Choices that are inconsistent with a stated valuation are thought to occur because of unspecified noise that corrupts the neural representation of value. Assuming that the noise is uncorrelated over time, the pattern of choices and response times in value-based decisions are modeled within the framework of Bounded Evidence Accumulation (BEA), similar to that used in perceptual decision-making. In BEA, noisy evidence samples accumulate over time until the accumulated evidence for one of the options reaches a threshold. Here, we argue that the assumption of temporally uncorrelated noise, while reasonable for perceptual decisions, is not reasonable for value-based decisions. Subjective values depend on the internal state of the decision maker, including their desires, needs, priorities, attentional state, and goals. These internal states may change over time, or undergo revaluation, as will the subjective values. We reasoned that these hypothetical value changes should be detectable in the pattern of choices made over a sequence of decisions. We reanalyzed data from a well-studied task in which participants were presented with pairs of snacks and asked to choose the one they preferred. Using a novel algorithm (Reval), we show that the subjective value of the items changes significantly during a short experimental session (about 1 hour). Values derived with Reval explain choice and response time better than explicitly stated values. They also better explain the BOLD signal in the ventromedial prefrontal cortex, known to represent the value of decision alternatives. Revaluation is also observed in a BEA model in which successive evidence samples are not assumed to be independent. We argue that revaluation is a consequence of the process by which values are constructed during deliberation to resolve preference choices.

Introduction

A central idea in decision theory and economics is that each good can be assigned a scalar utility value that reflects its desirability. The concept of utility, or subjective value, provides a common currency for comparing dissimilar goods (e.g., pears and apples) such that decision-making can be reduced to estimating the utility of each good and comparing them (von Neumann and Morgenstern, 1944; Samuelson, 1937; Montague and Berns, 2002). The idea is supported by studies that have identified neurons that correlate with the subjective value of alternatives in various brain structures, most notably the ventromedial prefrontal cortex, and it is so pervasive that decisions based on preferences are often referred to as “value-based decisions” (Kable and Glimcher, 2007; Kim et al., 2008; Padoa-Schioppa and Assad, 2006).

Choice and response time (RT) in simple perceptual and mnemonic decisions are often modeled within the framework of bounded evidence accumulation (BEA). The framework posits that evidence samples for and against the different options are accumulated over time until the accumulated evidence for one of the options reaches a threshold or bound (Ratcliff, 1978; Gold and Shadlen, 2007). A case in point is the random dot motion (RDM) discrimination task, in which participants must decide whether randomly moving dots have net rightward or leftward motion, while the experimenter controls the proportion of dots moving coherently in one direction, termed the motion strength (e.g., Gold and Shadlen, 2007). BEA models explain the choice, RT, and confidence in the RDM task under the assumption that the rate of accumulation, often termed the drift rate, depends on motion strength (van Den Berg et al., 2016; Kiani et al., 2014). Value-based decisions have also been modeled within the framework of BEA. The key assumption is that at any given time, decision-makers only have access to a noisy representation of the subjective value of each item, and the drift rate depends on the difference between the subjective values of the items (Krajbich et al., 2010; Thomas et al., 2019; Sepulveda et al., 2020; Bakkour et al., 2019).

A condition that renders the BEA framework normative is that the noise corrupting the evidence samples is independent, or equivalently, that the evidence samples are conditionally independent given the drift rate. For example, in modeling the RDM and other perceptual decision making tasks, evidence samples are assumed to be independent of each other, conditioned on motion strength and direction (e.g., Zylberberg et al., 2016). This assumption is sensible because (i) the main source of stochasticity in perceptual decision making is the noise affecting the sensory representation of the evidence, which has a short-lived autocorrelation, and (ii) these decisions are often based on an evidence stream (e.g., a dynamic random dot display) that provides conditionally independent samples, by design. The assumption of conditional independence justifies the process of evidence accumulation, because accumulation (or averaging) can only remove the noise components that are not shared by the evidence samples.

For value-based decisions, the assumption of conditional independence is questionable. Alternatives often differ across multiple attributes (e.g., Busemeyer and Townsend, 1993; Tversky, 1977). For example, when choosing between different snacks, they may differ in calories, healthiness, palatability, and so on (Suzuki et al., 2017). The weight given to each attribute depends on the decision-maker’s internal state (Noguchi and Stewart, 2018; Juechems and Summerfield, 2019). This internal state includes desires, needs, priorities, attentional state and goals. We use the term mindset, or state of mind, to refer to all of these internal influences on valuation. A mindset can be persistent. For example, a famished decision-maker may prioritize the nutritional content of each food when making a choice. Under less pressing circumstances, the salience of an attribute may be suggested by snack alternatives themselves. For example, seeing French fries may make us aware that we crave something salty, and saltiness becomes a relevant attribute informing the current decision and possibly future decisions too. The examples illustrate how a decision-maker’s mindset can shift rapidly or meander, based on the attributes in focus or the identity of the items under consideration (Shadlen and Shohamy, 2016; Stewart et al., 2006). Importantly, mindset is dynamic. It can change abruptly, motivated by a thought in an earlier trial or by interoception during deliberation (e.g., thirst). Unlike perceptual decision-making, where the expectation of a sample of evidence is thought to be fixed, conditional on the stimulus, the expectation of the evidence bearing on preference is itself potentially dynamic.

We sought to test the notion that the desirability of an item changes as a result of the deliberation that leads to a choice. We hypothesized that if subjective values are dynamic, then value-based decisions should exhibit serial dependencies when multiple decisions are made in a sequence. A choice provides information not only about which option is preferred, but also about the decision maker’s mindset at the moment of the choice (e.g., whether they prioritize satiation or palatability). Therefore, a choice is informative about future choices because the decision maker’s mindset is likely to endure longer than a single decision, or even multiple decisions.

We reanalyzed data from Bakkour et al. (2019). Participants were presented with pairs of snacks and had to choose the one they preferred. This Food choice task has been used extensively to study the sequential sampling process underlying value-based decisions (e.g., Krajbich et al., 2010). Crucially, in the Bakkour et al. (2019) experiment, each item was presented multiple times, allowing us to infer how preference for an item changes during a single experimental session. Using a novel algorithm we call Reval, we show that the subjective value of items changed over the session. The revaluation was replicated in a sequential sampling model in which successive samples of evidence are not assumed to be conditionally independent. We argue that the revaluation process we observed reflects a process by which the value of the alternatives is constructed during deliberation by querying memory and prospecting for evidence that bears on desirability (Lichtenstein and Slovic, 2006; Johnson et al., 2007).

Results

Food choice task

We re-examined data from a previous study in which 30 participants completed a food choice task (Bakkour et al., 2019). Prior to the main experiment, participants were asked to indicate their willingness to pay for each of 60 snack items on a scale from 0 to US$3 (Fig. 1A). We refer to these explicitly reported values as s-values, or vs (where s stands for ‘static’ as opposed to the ‘dynamic’ values we define below). In the main experiment (conducted in an MRI scanner), participants were shown pairs of images of previously rated snack items and had to choose which snack they would prefer to consume at the end of the study (Fig. 1B).

Food choice task

(A) In an initial ‘ratings’ task, participants were shown 60 individual appetizing snack items and asked to indicate how much they would be willing to pay for each item using a monetary scale ranging from $0 to $3.

(B) In the main experiment, participants were presented with pairs of snack items and asked to choose which one they would prefer to consume at the end of the session. After making their choice, the chosen item was highlighted by a square box for an additional 0.5 s. Each of the 30 participants completed 210 trials, with each item appearing 7 times during the experiment. A subset of 60 item pairs were repeated once.

(C) Proportion of trials in which participants selected the right item as a function of the difference in value between the right and left items (Δvs). Proportions were first determined for each participant and then averaged across participants. Error bars indicate the standard error of the mean (s.e.m.) across participants.

(D) Mean response time as a function of the difference in value between the right and left items. Error bars indicate the s.e.m. across participants. Red curves in panels C-D are fits of a drift-diffusion model (DDM).

The data from Bakkour et al. (2019) replicate the behavior typically observed in the task. Both choice and response time were systematically related to the difference in s-value, (Δvs), between the right and left items. Participants were more likely to choose the item to which they assigned a higher value during the rating phase (p<0.0001; ℋ0 : β1 = 0; Eq. 2). They were also more likely to respond faster when the absolute value of the difference between the items was greater (p<0.0001; ℋ0 : β1 = 0; Eq. 3).

The relationship between Δvs, choice, and response time is well described by a bounded evidence accumulation model (Krajbich et al., 2010; Bakkour et al., 2019). The solid lines in Fig. 1C-D illustrate the fit of such a model in which the drift rate depends on Δvs. Overall, the behavior of our participants in the task is similar to that observed in other studies using the same task (e.g., Krajbich et al., 2010; Folke et al., 2016; Sepulveda et al., 2020).

Limited power of explicit reports of value to explain binary choices

An intriguing aspect of the decision process in the food choice task is its highly stochastic nature. This is evident from the shallowness of the choice function (Fig. 1C): participants chose the item with a higher s-value in only 64% of the trials. This variability is typically attributed to unspecified noise when recalling item values from memory (e.g., Krajbich et al., 2010). An alternative explanation is rooted in constructive value theories, which suggest that the value of each item is constructed, not retrieved, during the decision process (Lichtenstein and Slovic, 2006; Shadlen and Shohamy, 2016; Johnson et al., 2007). This construction process is sensitive to the context in which it is elicited (e.g., the identity of items being compared), so the values reported during the valuation process may differ from those used in the choice task. According to this idea, the apparently stochastic choice is a veridical reflection of the constructed values.

If this were true, then the choice on any one cynosure trial—that is, the trial we are scrutinizing—would be better explained by values inferred from the choices on the other trials than by the s-values. We therefore compared two regression models that produce the log odds of the choice on each cynosure trial. The first regression model uses the s-values plus a potential bias for the left or right item. The second regression model includes one regression coefficient per item plus a left/right bias. It uses all the other trials (except repetitions of the identical pair of items) to establish the weights. While this model has more free parameters, the comparison is valid because we are using the models to predict the choices made on trials that were not used for model fitting. The better model is the one that produces larger log odds of the choice on the cynosure trial. As shown in Fig. 2, the second regression model is superior.

Individual choices are better explained by values inferred from the other trials than values reported in the ratings task

Gray data points represent the total log-likelihood of each participant’s choices, given two types of predictions: (abscissa) from a logistic regression, fit to the static values; (ordinate) from a procedure that infers the values based on choices on the other trials. Predictions derived from the other trials is better in all but four participants. The red markers were obtained using the same procedure, applied to choices simulated under the assumption that the s-values are the true values of the items. It shows that the inferential procedure is not guaranteed to improve predictions.

To ensure that this result is not produced artifactually from the algorithm, we performed the same analysis on simulated data. We fit the experimentally observed choices using a logistic regression model with Δvs and an offset as independent variables, and simulated the choices by sampling from Bernoulli distributions with parameter, p, specified by the logistic function that best fit each participant’s choices (i.e., weighted-coin flips). We repeated the model comparison using the simulated choices and found that, contrary to what we observed in the experimental data, the model using explicit value reports is the better predictor (Fig. 2, red).

Taken together, these analyses show that explicit value reports have limited power to predict choices, which partially explains their apparent stochasticity (Konovalov and Krajbich, 2019; Verhoef and Franses, 2003; Wardman, 1988). In the following sections, we elaborate on this observation. Not only do the values used to make the binary choices differ from the s-values, they drift apart during the experiment.

We show that these changes arise through the deliberative process leading to the preference decisions themselves.

Preferences change over the course of the experiment

In the experiment, a subset of the 60 snack pairs were presented twice, in a random order within the sequence of trials. These trials allow us to assess whether preferences change over the course of a session. For these duplicated item pairs, we calculate the average number of times that the same item was chosen on both presentations—which we refer to as the match probability. Participants were more likely to select the same option when presentations of the same pair were closer in time (Fig. 3). To assess the significance of this effect, we fit a logistic regression model using all pairs of trials with identical stimuli to predict the probability that the same item would be chosen on both occasions. The regression coefficient associated with the number of trials between repetitions was negative and highly significant (p<0.0001; t-test, Eq. 8). It therefore follows that preferences are not fixed, not even over the course of a single experimental session.

Preferences change over time

Probability of making the same choice on the two trials with the same item pair, shown as a function of the difference in trial number between them (Δtr). Trial pairs with identical items (N=1726) were sorted by Δtr, and the match probabilities were smoothed with a boxcar function with a width of 100 observations.

Choice alternatives undergo revaluation

We propose a simple algorithm to characterize how preferences changed over the course of the session. It assumes that on each decision, the value of the chosen item increases by an amount equal to δ, and the value of the unchosen item decreases by the same amount (Fig. 4A). We refer to the updated values as d-values, or vd, where d stands for ‘dynamic’.

Revaluation algorithm

(A) Schematic example of the revaluation algorithm applied to one decision. After a choice between items A and B, the value of the chosen item is increased by δ and the value of the unchosen item is decreased by the same amount. (B) Example of value changes due to revaluation, for three items, as a function of the presentation number within the session. In the experiment, each item was presented 7 times. (C) For a representative participant, deviance of the logistic regression model that uses the revalued values to explain the choices, for different values of δ. The best fitting value is ∼$0.15. The inset shows a histogram of the best-fitting δ values across participants.

Fig. 4B illustrates how the value of the items changes over the course of the session, for a given value of δ, for three snack items. For example, while the item shown with the green curve is initially very valuable, as indicated by its high initial rating, its value decreases over the course of the session each time it was not selected.

We determined the degree of revaluation that best explained the participants’ choices. For each participant, we find the value of δ that minimizes the deviance of a logistic regression model that uses the d-values to fit the choices made on each trial, where pchoice is the probability of choosing the item that was presented on the right. The d-values are initialized to the explicitly reported values for all items, and they are updated by plus or minus δ when an item is chosen or rejected, respectively. Importantly, the updated values only affect future decisions involving the items.

Fig. 4C shows the deviance of the logistic regression model for a representative participant, as a function of δ. For this participant, the best explanation of the choices is obtained with a value of δ ≈ $0.15. We fit the value of δ independently for each participant to minimize the deviance of the logistic regression model fit to the choices. On average, each choice changed the value of the chosen and unchosen items by $0.18 ± 0.016 (mean ± s.e.m., Fig. 4C, inset).

The values derived from the Reval algorithm explain the choices better than the explicit value reports. The choices are more sensitive to variation in Δvd, evidenced by the steeper slope (Fig. 5A). When Δvd and Δvs are allowed to compete for the same binomial variance, the former explains away the latter. This assertion is supported by a logistic regression model that incorporates both Δvs and Δvd as explanatory variables (Eq. 7). The coefficient associated with Δvs is not significantly different from zero while the one associated with Δvd remains positive and highly significant (Figure 5–Figure Supplement 1).

Revaluation explains choice and RT better than explicit values

(A) Proportion of rightward choices (top) and mean response time (bottom) as function of the difference in d-value between the two items. The red solid lines are fits of a drift-diffusion model that uses the d-values. The dashed line corresponds to the fits of a DDM that uses the s-values (same as in Fig. 1C-D). Error bars indicate s.e.m. across trials. Participants are more sensitive to d-values than s-values (top) and the d-values better explain the full range of RTs (bottom). (B) Percentage of variance in response times explained by a DDM in which the drift rate depends on either Δvd (abscissa) or Δvs (ordinate). Each data point corresponds to a different participant. For most participants, the model based on the dynamic values explained a greater proportion of the variance. (C) d-values are better than s-values at predicting the difficulty of a decision as reflected in the response times. Data points represent the difference in mean RTs between difficult and easy decisions. Positive values indicate that difficult decisions take longer on average than easy ones. Difficult and easy are defined relative to the median of the absolute value of Δvs (left) or Δvd (right). The lines connect the mean RTs of each participant. P-value is from a paired t-test.

Figure 5–Figure supplement 1. Dynamic values explain away the effect of static values on choice.

Figure 5–Figure supplement 2. Comparison of DDM fits using static and dynamic values.

Figure 5–Figure supplement 3. Similar δ values obtained by Reval and logistic regression.

More surprisingly, Reval allows us to explain the response times better than the explicit value reports, even though RTs were not used to establish the d-values. We used the d-values to fit a drift-diffusion model to the single-trial choice and response time data, and compared this model with the one that was fit using the s-values (Fig. 5A). To calculate the fraction of RT variance explained by each model, we subtracted from each trial’s RT the models’ expectation, conditional on Δvx (with x ∈ {e, r}) and choice. The model that relies on the d-values explains a larger fraction of variance in RT than the model that relies on the s-values (Fig. 5B). This indicates that the re-assignment of values following Reval improved the capacity of a DDM to explain the response times.

The DDM that uses the dynamic values also explains the combined choice-RT data better than the one that uses the static values. We compared their goodness of fit using the Bayesian Information Criteria (BIC), penalizing the DDM that uses the dynamic values for the revaluation update parameter, δ. For all participants, the DDM that uses the dynamic values provided a better fit than the DDM that uses the static values (Figure 5–Figure Supplement 2A). To control for the possibility that the model comparison is biased by the extra parameter in the dynamic model (δ), we simulated choice and RT data for each participant from the DDM model fit to the static values, and fit these simulated data to the DDMs using static and dynamic values (in the latter case applying the Reval algorithm prior to fitting). For the simulated data, the model comparison favored the DDM using static values for most participants (Figure 5–Figure Supplement 2B), indicating that the additional parameter in the dynamic model does not strongly bias the model comparison.

The time it takes to make a decision, and the difference in value between the items under comparison, can be considered complementary measures of decision difficulty. On average, the more similar in value the two items are, the longer it would take to commit to a choice. Under this assumption, we can compare how well the static and the dynamic values predict the difficulty of the choices as judged by their response times. The application of Reval revealed that some decisions that were initially considered difficult, because Δvs was small, were actually easy, because Δvd was large, and vice versa. Grouping trials by the Δvd led to a wider range of mean RTs compared to when we grouped them by Δvs (Fig. 5C). The effect can also be observed for individual participants. For each participant, we grouped trials into two categories depending on whether the difference in value was less than or greater than the median difference. We then calculated the mean RT for each of the two groups of trials. The difference in RT between the two groups was greater when we grouped the trials using the d-values than when we used the s-values. This implies the d-values were better than the s-values at assessing the difficulty of a decision as reflected in the response time.

We verified that the improvement in fit was not just due to the additional free parameter (δ). To do this, we again used simulated choices sampled from logistic regression models fit to the participants’ choices, as we did for Fig. 2. Because the choices are sampled from logistic functions fit to the choice data, they lead to a psychometric function that is similar to that obtained with the experimental data. We reasoned that if revaluation were an artifact of the analysis method, then applying the revaluation algorithm to these simulated data should lead to values of δ and goodness of fit similar to those of the real data. To the contrary, (i) the optimal values of δ for the simulated data were close to zero (Fig. 6A); (ii) the reduction in deviance after applying Reval was negligible compared to the reduction in the actual data (Fig. 6B); and (iii) we found no difference in the RT median splits between s-values and d-values (Fig. 6C). This shows that the improvements in fit quality due to Reval are neither guaranteed nor an artifact of the procedure.

No revaluation in simulated data

(A) Histogram of the best-fitting revaluation update (δ) for data simulated by sampling choices from a logistic function fit to the pa best-fitting δ values for the simulated choices are centered around 0. For reference, we have also included a histogram of the δ values obtained from the fits to the participants’ data, showing all positive values (gray). (B) Deviance of the logistic regression model used to explain the choices (Eq. 1), fit using either the static values (ordinate) or the Reval algorithm (abscissa). Each data point corresponds to a different participant. Experimental data are shown in gray and simulated data (as in panel A) are shown in red. The marked reduction in deviance in the experimental data is absent in the data simulated by sampling from logistic regressions fit to the static values. (C) Similar to Fig. 5C, for the simulated data. The values obtained from Reval were no better than the static values at explaining the RTs, as expected, since the δ values were ∼0 and thus vd ≈ vs.

Imperfect value reports do not explain revaluation away

The idea that a choice can induce a change in preference is certainly not new (Festinger, 1957). Choice-induced preference change (CIPC) has been documented using a free-choice paradigm (Brehm, 1956), whereby participants first rate several items, and then choose between pairs of items to which they have assigned the same rating, and finally rate the items again. A robust finding is that items that were chosen are given higher ratings and items that were not chosen are given lower ratings relative to pre-choice ratings, leading to the interpretation that the act of choosing changes the preferences for the items under comparison. However, it has been suggested that the CIPC demonstrated with the free-choice paradigm can be explained as an artifact (Chen and Risen, 2010). Put simply, the initial report of value may be a noisy rendering of the true latent value of the item. If two items, A and B, received the same rating but A was chosen over B, then it is likely that the true value for item A is greater than for item B, not because the act of choosing changes preferences, but because the choices are informative about the true values of the items, which are unchanging.

We examined whether Reval could be explained by the same artifact. We considered the possibility that the items’ valuation in the choice phase are static but potentially different from those reported in the ratings phase. If the values are static, but different from those explicitly reported, then Reval could still improve choice and RT predictions by revealing the true subjective value of the items.

We reasoned that if values were static, the improvements we observed in the logistic fits when we applied Reval should be the same regardless of how we ordered the trials before applying it. To test this, we applied Reval in the direction in which the trials were presented in the experiment, and also in the reverse direction (i.e., from the last trial to the first). If the values were static, then the quality of the fits should be statistically identical in both cases. In contrast, we observed that the variance explained by Reval was greater (i.e., the deviance was lower) when it was applied in the correct order than when it was applied in the opposite order (Fig. 7A; p<0.0001, paired t-test). This rules out the possibility that the values were static. Moreover, the values produced by applying Reval in the reverse direction explained the choices better than the static values (Fig. 7B). This might seem counterintuitive, given that the initial values for the Reval algorithm are the s-values, which are explicitly reported before the main experiment.

Reval is sensitive to trial order

(A) Deviance obtained by applying Reval to the trials in the order in which they were completed (abscissa) and in the reverse order (ordinate). Each data point corresponds to a different participant. The deviance is greater (i.e., the fits are worse) when Reval is applied in the reverse direction. (B) The deviance of the logistic regression model used to explain the choices (Eq. 1), obtained by applying Reval in the backward direction (ordinate), is lower than the deviance obtained using the static values (abscissa). Each data point corresponds to a different participant. Experimental data are shown in gray and data simulated from the logistic fits to the static values (as in Fig. 6A-B) are shown in red.

In a later section, we show that this effect stems from the same process that gives rise to revaluation (Is revaluation a byproduct of deliberation?).

Asymmetric value-updating for chosen and unchosen options

So far we have assumed that a choice increases the value of the chosen option by δ and decreases the value of the unchosen option by the same amount. Here, we evaluate the possibility that the degree of revaluation is different for the chosen and unchosen options. We fit a variant of the Reval algorithm with two values of δ, one for the chosen option (δchosen) and one for the unchosen option (δunchosen). Fig. 8 shows the values that best fit the data for each participant. For each participant, δchosen>0 and δunchosen<0; in other words, the value of the chosen item typically increases, while the value of the unchosen item tends to decrease following a choice. Further, for most participants, the degree of revaluation is greater for the chosen option than for the unchosen option. As we speculate in the discussion, this result may be related to the unequal distribution of attention between the chosen and unchosen items (Krajbich et al., 2010).

Stronger revaluation for the chosen than for the unchosen item

We fit a variant of the Reval algorithm that includes separate update values (δs) for the chosen and unchosen options. The best-fitting δ value for the chosen option (abscissa) is plotted against the best-fitting value for the unchosen option (ordinate). Each data point corresponds to one participant. The increase in value for the chosen option is greater than the decrease in value for the unchosen option (paired t-test).

Representation of revalued values in the ventromedial prefrontal cortex

Several brain areas, in particular the ventromedial prefrontal cortex (vmPFC), have been shown to represent the value of decision alternatives during value-based decisions (Kennerley et al., 2009; Plassmann et al., 2007; Bartra et al., 2013). Based on our finding that the d-values provide a better explanation of the behavioral data than the s-values, we reasoned that the d-values might explain the BOLD activity in these areas beyond that explained by the s-values. We included both the s-value and the d-value of the chosen item in a whole-brain regression analysis of BOLD activity. This parameterization reveals significant correlation of the BOLD signal in the vmPFC with d-value, controlling for s-value (Fig. 9 and Table S1). In fact, in a separate model that only included s-value, the effect of s-value on BOLD in the vmPFC did not survive correction for familywise error rate at a whole-brain level (Figure 9–Figure Supplement 1 and Table S2 top). In contrast, another model that only included d-value revealed a robust effect of d-value on BOLD in vmPFC that survived whole-brain correction (Figure 9–Figure Supplement 1 and Table S2 middle). Finally, to evaluate whether the effect shown in Fig. 9 is not simply captured by the difference in d-value and s-value, we ran a fourth model that included only (d-value − s-value). The effect of this difference between d-value and s-value on BOLD in vmPFC did not survive whole-brain correction (Figure 9–Figure Supplement 1 and Table S2 bottom). Collectively, these findings provide additional evidence for revaluation, as capturing a meaningful aspect of the data, in the sense that it accounts for the activity of brain areas known to reflect the value of the choice alternatives.

Revaluation reflected in BOLD activity in ventromedial prefrontal cortex

Brain-wide fMRI analysis revealed a significant correlation between d-values and activity in the vmPFC, after controlling for s-values. The statistical map was projected onto the cortical surface. Shown here are the medial view of the right and left hemispheres of a semi-inflated surface of a template brain. Heatmap color bars range from z-stat = 3.1 to 3.6. The map was cluster corrected for familywise error rate at a whole-brain level with an uncorrected cluster-forming threshold of z = 3.1 and corrected extent of p < 0.05. The full unthresholded map can be viewed here: https://identifiers.org/neurovault.image:869963.

Figure 9–Figure supplement 1. Representation of d-value, s-value and their difference in BOLD activity.

Revaluation in other datasets of the food-choice task

To assess the generality of our behavioral results, we applied Reval to other publicly available datasets. All involve binary choices between food snacks, similar to Bakkour et al. (2019). We analyze data from experiments reported in Folke et al. (2016) and from the two value-based decision tasks reported in Sepulveda et al. (2020).

Reval yields results that are largely similar to those observed in the data from Bakkour et al. (2019). The values derived from Reval led to a better classification of choice difficulty than the explicit value reports (Fig. 10A). In all three datasets, the δ values were significantly larger than those obtained from simulated data under the assumption that the values were static and equal to the explicitly reported values (Fig. 10B). Furthermore, the reduction in the deviance resulting from the application of Reval (Eq. 7) was significantly greater than the reduction observed in simulated data (Fig. 10C). [All p-values, derived from two-tailed paired t-tests, are shown in the figure].

Revaluation observed in other datasets

We applied the Reval method to other publicly available datasets of the food choice task. In the experiment of Folke et al. (2016) (first column), participants reported their willingness to pay (WTP) for each of 16 common snack items. In the choice task, they were presented with each unique pair of items and asked to choose the preferred item. Each unique pair was presented twice for a total of 240 trials per participant. In the experiment of Sepulveda et al. (2020) (second and third columns), participants (N=31) reported their willingness to pay (WTP) for each of 60 snack items. They were then presented with pairs of items from which to choose. Pairs were selected based on participants’ WTP reports to provide comparisons between pairs of high-value, low-value and mixed-value items. The choice task was performed under two framing conditions: like-framing, selecting the more preferred item, and dislike framing, selecting the less preferred item. The task consisted of six alternating blocks of like- and dislike-framing (40 trials per block). (A) RT difference between easy and difficult trials, determined as a median split of |Δv|. Same analysis as in Fig. 5C. (B) Histogram of the best-fitting revaluation update (δ) for data simulated by sampling choices from a logistic function fit to the participant’s choices (red), and for the actual data (gray). Same analysis as in Fig. 6A. (C) Comparison of the deviance with and without Reval. Same analysis as in Fig. 6A. (D) Comparison of the deviance applying Reval in the forward and backward directions. Same analysis as in Fig. 7A. All p-values shown in the figure are from paired t-tests.

In the dataset from Folke et al. (2016), the deviance was significantly smaller when Reval was applied in the forward than in the backward direction, replicating the result in our main experiment. However, in the dataset of Sepulveda et al. (2020), no significant difference in deviance was observed (Fig. 10D). We do not know what explains this discrepancy, although we believe that the differences in experimental design may play a role. In the experiment of Sepulveda et al. (2020), unlike the other two datasets that we analyzed, participants performed the experiment in two framing conditions: one in which they chose the item the liked the most, and another one in which the chose the item the disliked the most. These two conditions alternated in short blocks of 40 trials. This alternation may affect valuation in a way that is not captured by the Reval algorithm. We expand on this in Discussion.

Is revaluation a byproduct of deliberation?

We hypothesize that the sequential dependencies we identified with Reval may be a corollary of the process by which values are constructed during deliberation. The subjective value of an item depends on the decision-maker’s mindset, which may change more slowly than the rate of trial presentations. Therefore, the subjective value of an item on a given trial may be informative about the value of the item the next time it is presented. Subjective values are not directly observable, but choices are informative about the items’ value.

We assessed the plausibility of this hypothesis with a bounded evidence accumulation model that includes a parameter that controls the correlation between successive evidence samples for a given item. We call this the correlated-evidence drift-diffusion model (ceDDM). We assume that the decision is resolved by accumulating evidence for and against the different alternatives until a decision threshold is crossed.

The model differs from standard drift-diffusion, where the momentary evidence is a sample drawn from a Normal distribution with expectation equal to Δvs plus unbiased noise, . Instead, the value of each of the items evolves separately such that the expectations of its value updates are constructed as a Markov chain Monte Carlo (MCMC) process thereby introducing autocorrelation between successive samples of the unbiased noise (see Methods). Crucially, the correlation is not limited to the duration of a trial but extends across trials containing the same item. When an item is repeated in another trial, the process continues to evolve from its value at the time a decision was last made for or against the item.

We fit the model to the data from Bakkour et al. (2019). The model was able to capture the relationship between choice, response time and Δvs (Fig. 11A). Fig. 11B shows the degree of correlation in the evidence stream as a function of time, for the model that best fit each participant’s data. After 1 second of evidence sampling, the correlation was 0.1062 ± 0.0113 (mean ± s.e.m. across participants). This is neither negligible (which would make the model equivalent to the DDM) nor very high (which would render sequential sampling useless, since it can only average out the noise that is not shared across time).

Revaluation occurs in a DDM with temporally-correlated noise

A drift-diffusion model with non-independent noise (ceDDM) captures the main features of revaluation. (A) The ceDDM accounts for choices (top) and response times (bottom), plotted as a function of the difference in values obtained from explicit reports (Δvs). Same data as in Fig. 1C-D. Red curves are simulations of the best-fitting model. Each trial was simulated 100 times. Simulations were first averaged within trials and then averaged across trials. Error bars and bands indicate s.e.m. across trials. (B) Noise correlations as a function of time lag, obtained from the best-fitting model. Each curve corresponds to a different participant. Each curve corresponds to a different participant. (C) δ parameters derived by applying Reval to simulated data from the best fitting ceDDM model to each participant’s data. As in the data, δ > 0 for all participants. (D) Similar analysis as in Fig. 5C applied to simulations of the ceDDM. As for the data, Reval increased the range of RTs obtained after grouping trials by difficulty (by s-values on the left and d-values on the right; p-value from paired t-test). (E) Similar analysis to that of Fig. 7A, using the simulated data. As observed in the data, the deviance resulting from applying Reval in the correct trial order (abscissa) is smaller than when applied in the opposite order (p-value from paired t-test). (F) Similar analysis to that of Fig. 7B, using the simulated data.

The assumptions embodied by the ceDDM are consistent with the results of the Reval analysis. We applied the Reval algorithm to simulated data obtained from the best-fitting ceDDM. The results were in good agreement with the experimental data. The best-fitting δ values were positive for all participants and in a range similar to what we observed in the data (Fig. 11C). Reval increased the range of RTs when trials were divided by difficulty, implying that Reval led to a better classification of easy and difficult decisions (Fig. 11D). Reval applied to the trials in the true order explained the simulated choices better than when applied in the opposite direction (Fig. 11E). This is because the model assumes that when an item first appears, the last sample obtained for that item was the value reported in the ratings phase for that item. As more samples are obtained for a given item, the correlation with the explicit values gradually decreases. Additionally, the values obtained from applying Reval in the backward direction provided a better explanation of the simulated choices than the static values (Fig. 11F), mirroring the pattern observed observed in the behavioral data (Fig. 7B). Taken together, the success of ceDDM implies that the sequential dependencies we identify with Reval may be the result of a value construction process necessary to make a preferential choice.

Discussion

Sequential dependencies and choice-induced preference change

We identified sequential dependencies between choices in a value-based decision task. Participants performed a task in which they had to make a sequence of choices among a limited set of items. The best explanation for future choices was obtained by assuming that the subjective value of the chosen item increases and the value of the unchosen item decreases after each decision. Evidence for revaluation was obtained by analyzing the probability that participants make the same decision in pairs of trials with identical options. We also identified revaluation using an algorithm we call Reval. The same algorithm allowed us to identify revaluation in other datasets obtained with the food-choice task (Folke et al., 2016; Sepulveda et al., 2020).

The sequential effects we identified can be interpreted as a manifestation of choice-induced preference change. The usual paradigms for detecting the presence of CIPCs are based on the comparison of value ratings reported before and after a choice (for a review see Izuma and Murayama, 2013; Enisman et al., 2021). After a difficult decision, the rating of the chosen alternative often increases and that of the rejected alternative often decreases—an effect termed the “spreading of alternatives”. Many variants of the free choice paradigm have been developed to control for or eliminate the statistical artifact reported by Chen and Risen (2010). One common approach is to compare the “spreading of alternatives” observed in the free-choice paradigm (rate-choose-rate, or RCR) with a control task in which a different set of participants rate the items twice before the choice phase (RRC). Any spread observed in the RRC condition cannot be explained by the CIPC, since in the RRC condition there is no choice between the two rating phases. The CIPC is measured indirectly, as the difference in the spread of the alternatives between the RCR and the RRC. Other approaches involve asking participants to rate an item that they are led to believe they have chosen, when in fact they have not (Sharot et al., 2010; Johansson et al.,2014). Any change in ratings cannot be due to the information provided by a choice, since no real choice was made. In addition to the complications introduced by deceiving the participants (e.g., participants may suspect the deception but not mention it to the experimenter), the elimination of a real choice prevents these paradigms from being used to study the process through which subjective values undergo revision during decision formation.

In contrast, our approach to identify changes in value does not require pre- and post-choice ratings. Instead, it requires a sequence of trials in which the same items are presented multiple times (as in Luettgau et al., 2020). The revaluation effect we find cannot be explained by the artifact identified by Chen and Risen (2010). Using trials with identical items, we show that the nearer in time the trials with identical items are to each other, the more likely people are to choose the same option. Further, the revaluation algorithm explains choices better when applied in the order in which the trials were presented than when applied in the reverse order. These observations are inconsistent with the notion that item values are fixed (i.e., do not change) during the experiment, regardless of whether values are the same or different from those reported during the rating phase.

Revaluation during of after deliberation?

We cannot determine with certainty whether the revaluation occurs after the decision or during the deliberation process leading up to the decision. At face value, it might seem that Reval implements change after each decision (Festinger, 1957). Yet, Reval simply identifies a change in value, which may well occur during the deliberation leading to the decision, perhaps owing to a comparison of other items (on other trials) that happen to suggest a dimension of comparison that increases in importance on the current trial (Lee and Daunizeau, 2020; Lichtenstein and Slovic, 2006). More broadly, the subjective value of an option depends on the mindset of the decision maker. This internal state, which in the food-choice task includes aspects such as degree of satiety or sugar craving, can vary over time, causing the value of the items to vary as well. If changes in mindset are slow—that is, lasting longer than the duration of a decision—then the value of items will be correlated over time.

We proposed a decision model (ceDDM) in which evidence samples are correlated over time. Fitting the model to account for each participant’s choices and response times produces a revaluation of magnitude similar to what we observed experimentally. It also predicts that applying Reval in the direction in which the trials were presented explains the choices better than applying it in the opposite direction, as we observed in the data. This modeling exercise suggests that the CIPC-like effects we identified may be due to processes that occur during the deliberation leading up to a choice, rather than post-decision processes that attempt to reduce cognitive dissonance. To be clear, we interpret the ceDDM only as a proxy for a variety of more nuanced processes. If the mindset endures many individual decisions, the subjective value of an item will be correlated over time. While the ceDDM captures only a small aspect of this complex process, it has allowed us to explain the sequential dependencies we identified with Reval.

The ceDDM belongs to a class of sequential sampling models in which the drift rate varies over time. Such models have already been studied in the context of value-based decisions. For example, in the attentional drift-diffusion model (Krajbich et al., 2010), the drift rate varies depending on which item is attended, as if the value of the unattended items are discounted by a multiplicative factor. In Dynamic Field Theory (Busemeyer and Townsend, 1993), the drift rate varies depending on which attribute is attended. Recently, Lee and Pezzulo (2022) showed that a sequential sampling model in which the drift rate varies over time can explain the ‘spreading of alternatives’ (SoA) characteristic of choice-induced preference change. Lee and Pezzulo (2022) propose that the initial rating of the items may be constructed using only the most salient attributes of each item, while in a difficult decision more attributes may be considered, leading to a revaluation that informs the rating reported after the decision phase (see also Voigt et al., 2019). Consistent with our proposal, Lee and Pezzulo (2022) argue that thinking about non-prominent features during decision-making increases the likelihood that these features will be recalled when evaluating options in subsequent instances.

More revaluation for the chosen than the unchosen item

We observed that the degree of revaluation was higher for the chosen item than for the unchosen item. This was revealed by a variant of the Reval algorithm in which we allowed both items to have different updates. We speculate that this difference can be explained by the asymmetric distribution of attention between the chosen and unchosen items. It is known that the chosen item is looked at longer than the unchosen item (Krajbich et al., 2010). Further, CIPC is more likely for items that are remembered to have been chosen or unchosen (Salti et al., 2014). So one possibility is that the revaluation is larger for the chosen than for the unchosen item because participants spent more time looking at the chosen item and thus are more likely to remember it, leading to a larger change in value (Voigt et al., 2019).

Another possibility derives from the constructive view of preferences and the potential role of attention in decision-making. It is often assumed that value-based decisions involve gathering evidence from different alternatives, and that more evidence is gathered from alternatives that are attended to for longer (Callaway et al., 2021; Li and Ma, 2021; Krajbich et al., 2010). In the ceDDM, the correlation in value for a given item decreases with the number of evidence samples collected from the item (Fig. 11B). Therefore, the more that attention is focused on a given item, the greater the difference between the item’s value before and after the decision. Because chosen items are attended to for longer than unchosen items (e.g., Krajbich et al., 2010), the chosen item should exhibit larger revaluation than the unchosen one, which is what we observed in the data (Fig. 8).

Limitations of our study

One limitation of our study is that we only examined tasks in which static values were elicited from explicit reports of the value of food items. It remains to be determined if other ways of eliciting subjective values (e.g., Jensen and Miller, 2010) would lead to similar results. We think so, as the analysis of trials with identical item pairs (Fig. 3) and the difference between forward and backward Reval (Fig. 7A) are inconsistent with the notion that values are static, regardless of their precise value. It also remains to be determined if our results will generalize to non-food items whose value is less sensitive to satiety and other dynamic bodily states. Perceptual decisions also exhibit sequential dependencies, and it remains to be explored whether these can be explained as a process of value construction, similar to what we propose here for the food-choice task (Gupta et al., 2024; Cho et al., 2002; Zylberberg et al., 2018; Abrahamyan et al., 2016).

Another limitation of our study is that, in one of the datasets we analyzed (Sepulveda et al., 2020), applying Reval in the forward direction was no better than applying it in the backward direction (Fig. 10). We speculate that this failure is related to idiosyncrasies of the experimental design, in particular, the use of alternating blocks of trials with different instructions (select preferred vs. select non-preferred). More importantly, Reval applied in the backward direction led to a significant reduction in deviance relative to that obtained using the static values (Fig. 7B). This reduction was also observed in the ceDDM, suggesting that the effect may be explained by changes in valuation during deliberation. However, we cannot discard a contribution from other, non-dynamic changes in valuation between the rating and choice phase including contextual effects (Lichtenstein and Slovic, 2006), stochastic variability in explicit value reporting (Polanía et al., 2019), and the limited range of numerical scales used to report value.

Finally, we emphasize that the ceDDM should be interpreted as a proof-of-principle model used to illustrate how stochastic fluctuations in item desirability can explain many of our results. We chose to model value changes following an MCMC process. However, other stochastic processes or other ways of introducing sequential dependencies (e.g., variability in the starting point of evidence accumulation) may also explain the behavioral observations. Furthermore, there likely are other ways to induce changes in the value of items other than through past decisions. For example, attentional manipulations or other experiences (e.g., actual food consumption) may change one’s preference for an item. The current version of the ceDDM does not allow for these influences on value, but we see no fundamental limitation to incorporating them in future instantiations of the model.

Concluding remarks

Our research contributes to a growing body of work exploring the impact of memory on decision-making and preference formation (Biderman et al., 2020), and in particular to the CIPC. It has been suggested that the retrieval of an item’s value during decision-making renders it susceptible to modification, leading to a revaluation that influences subsequent valuations through a process that has a neural correlate in the hippocampus (Luettgau et al., 2020). The link between memorability and preference is also supported by experiments in which the presentation of an item coincides with an unrelated rapid motor response that increases subsequent preference for the item (Botvinik-Nezer et al., 2021) and by experiments demonstrating that people prefer items to which they have previously been exposed (Zajonc, 1968). As in these studies, ours also highlights the role of memory in revaluation. Due to the associative nature of memory, successive evidence samples are likely to be dependent (Rhodes and Turvey, 2007). A compelling illustration of this effect was provided by Elias Costa and colleagues (Elias Costa et al., 2009). Participants were asked to report the first word that came to mind when presented with a word generated by another participant, which was then shown to yet another participant. The resulting chain resembled Lévy flights in semantic space, characterized by mostly short transitions to nearby words and occasional large jumps. Similar dynamic processes have been used to describe eye movements during visual search (Bella-Fernández et al., 2021) and the movement of animals during reward foraging (Brown et al., 2007; Hills et al., 2015). It is intriguing to consider that a similar process may describe how decision-makers search their memory for evidence that bears on a decision.

Methods

Food choice task

A total of 30 participants completed the snack task, which consisted of a rating and a choice phase. The experimental procedures were approved by the Institutional Review Board (IRB) at Columbia University, and participants provided signed informed consent before participating in the study. The data were previously published in Bakkour et al. (2019).

Rating Phase

Participants were shown a series of snack items in a randomized order on a computer screen. They indicated their willingness to pay (WTP) by using the computer mouse to move a cursor along an analog scale ranging from $0 to $3 at the bottom of the screen. The process was self-paced, and each snack item was presented one at a time. After completing the ratings for all 60 items, participants were given the opportunity to revise their ratings. The 60 items were re-displayed in random order, with the original bids displayed below each item. Participants either chose to keep their original bid by clicking “NO” or to revise the bid by clicking “YES,” which re-displayed the analog scale for bid adjustment. We take the final WTP that is reported for each item as the corresponding explicit value (s-value).

Choice phase

From the 60 rated items, 150 unique pairs were formed, ensuring variation in Δvs. Each of the 60 items was included in five different pairs. The 60 item pairs were presented twice, resulting in a total of 210 trials per participant. Item pairs were presented in random order, with one item on each side of a central fixation cross. Participants were instructed to select their preferred food item and were informed that they would receive their chosen food from a randomly selected trial to consume at the end of the experiment. The task took place in an MRI scanner. Participants indicated their choice on each trial by pressing one of two buttons on an MRI-compatible button box. They had up to 3 seconds to make their choice. Once a choice was made, the chosen item was highlighted for 500 ms. Trials were separated by an inter-trial interval (ITI) drawn from a truncated exponential distribution with a minimum ITI of 1 and a maximum ITI of 12 seconds. The resulting distribution of ITIs across trials had a true mean of 3.05 seconds and a standard deviation of 2.0 seconds.

Data analysis

Association between the s-values, choice and RT

We used the following logistic regression model to evaluate the association between the s-values and the probability of choosing the item on the right: where Ii is an indicator variable that takes the value 1 if the trial was completed by subject i and 0 otherwise. We used a t-test to evaluate the hypothesis that the corresponding regression coefficient is zero, using the standard error of the estimated regression coefficient.

Similarly, we used a linear regression model to test the influence ofΔvs on response times: where | · | denotes absolute value and Σvs is the sum of the value of the two items presented on each trial. The last term was included to account for the potential influence of value sum on response time (Smith and Krajbich, 2019).

Predicting choices in cynosure trials

We used two logistic regression models to predict the choice in each trial using observations from the other trials. We refer to the trial under consideration as the cynosure trial (Fig. 2). One model uses the explicitly reported values: while the other model uses the choices made on other trials: where For this model, we included an L2 regularization with λ = 0.5. Both models were fit independently for each participant. We only included trials with the first appearance of each item pair (i.e., we did not include the repeated trials) so that the choice prediction for the cynosure trial is not influenced by the choice made in the paired trial containing the same items as in the cynosure trial.

Association between d-values and choice

We tested the association between d-values and choice with a logistic regression model fit to the choices. We included separate regressors for Δvd and Δvs: The model was fit separately for each participant. Figure 5–Figure Supplement 1 shows the regression coefficients associated with Δvs and Δvd.

Choice and response time functions

When plotting the psychometric and chronometric functions (e.g., Fig. 1C-D), we binned trials depending on the value of Δvs (or Δvd). The bins are defined by the following edges: { −∞,-1.5,-0.75,-0.375,-0.1875,-0.0625, 0.0625,0.1875,0.375,0.75,1.5,∞ }. We averaged the choice or RT for the trials (grouped across participants) within each bin and plotted them aligned to the mean Δvx of each bin.

Match probability

We used logistic regression to determine if the probability of giving the same response to the pair of trials with identical stimuli depended on the number of trials in between (Fig. 3). The model is: where pmatch is the probability of choosing the same item on both occasions, Ii is an indicator variable that takes a value of 1 if the pair of trials correspond to subject i, and zero otherwise, and T1st and T2nd are the trial number of the first and second occurrences of the same pair, respectively. We used a t-test to evaluate the hypothesis that β2 = 0 (i.e., that the separation between trials with identical stimuli had no effect on pmatch.

Drift-diffusion model

We fit the choice and RT data with a drift-diffusion model. It assumes that the decision variable, x, is given by the accumulation of signal and noise, where the signal is a function of the difference in value between the items, Δv, and the noise is equal to , where dt is the time step, such that the accumulated noise after 1 second of unbounded accumulation, the variance of the accumulated noise is equal to 1. The decision variable follows the difference equation, where ηt is sampled from a normal distribution with a mean 0 and variance 1, k is a signal-noise parameter, μ is the drift rate and μ0 is a bias coefficient that is included to account for potential asymmetries between right and left choices.

We assume that the drift rate is a (potentially nonlinear) function of Δvx. We parameterize this relationship as a power law, so that where sign is the sign operation, || indicates absolute value, and γ is a fit parameter.

The decision terminates when the accumulated evidence reaches an upper bound, signaling a rightward choice, or a lower bound, signaling a leftward choice. The bound is assumed to collapse over time. It is constant until time d, and then it collapses at rate a: Collapsing bounds are needed to explain why choices that are consistent with the value ratings are usually faster than inconsistent choices for the same Δvx.

The response time is the sum of the the decision time, given by the time taken by the diffusing particle to reach of the bounds, and a non-decision time which is assumed to be normally distributed with mean μnd and standard deviation σnd.

The model has 8 parameters: {k, B0, a, d, γ, μ0, μnd, σnd }. The standard deviation of the non-decision times (σnd) was fixed to 0.05 s. For the fits shown in Fig. 1C-D and Fig. 5A, we fit the model to grouped data from all participants. For the analysis of variance explained (Fig. 5) and model comparison (Figure 5–Figure Supplement 2), we fit the model separately for each participant. The model was fit to maximize the log of the likelihood of the parameters given the single-trial choice and RT: We evaluate the likelihood by numerically solving the Fokker-Planck (FP) equation that described the dynamics of the drift-diffusion process, using the Chang-Cooper fully-implicit method (Chang and Cooper, 1970; Kiani and Shadlen, 2009; Zylberberg et al., 2016). For computational considerations, we bin the values of Δvx to multiples of $0.1. From the numerical solution of the FP equation, we obtain the distribution of decision times, which is convolved with the truncated Gaussian distribution of non-decision latencies. The truncation ensures that the non-decision times are non-negative, which could otherwise occur during the optimization process for large values of σnd. The parameter search was performed using the Bayesian Adaptive Direct Search (BADS) algorithm (Acerbi and Ma, 2017).

Revaluation algorithm

The Reval algorithm was applied to each participant independently. The values are initialized to those reported during the ratings phase. They are then revised, based on the outcome of each trial, in the order of the experiment. The value of the chosen item is increased by δ and the value of the unchosen item is decreased by the same amount. The revaluation affects future decisions in which the same item is presented.

We searched for the value of δ* that minimizes the deviance of the logistic regression model specified by Eq. 1. The model’s deviance is given by: where the sum is over trials and ĉi is the probability assigned to the choice on trial i obtained from the best-fitting logistic regression model.

We complemented this iterative algorithm with a second approach that estimates δ* using the history of choices preceding each trial. Nearly identical δ values are derived using a single logistic regression model in which the binary choice made on each trial depends on the number of times each of the two items was selected and rejected on previous trials. The model is: where, as before, Ii is an indicator variable that takes a value of 1 if the trial was completed by subject

i and 0 otherwise. The key variable is Δch. It depends on the number of past trials in which the item presented on the right in the current trial was chosen and not chosen , and similarly, the number of past trials in which the item presented on the left in the current trial was chosen and not chosen: The variable Δch represents the influence of past choices. The signs in Eq. 15 are such that a positive (negative) value of Δch indicates a bias toward the right (left) item. To obtain the δ* in units equivalent to those derived with Reval, we need to divide the regression coefficient β2,i by the sensitivity coefficient β1,i, separately for each subject i. As can be seen in Figure 5–Figure Supplement 3, the values obtained with this method are almost identical to those obtained with the Reval algorithm.

Correlated-evidence DDM

The model assumes that at each moment during the decision-making process, the decision-maker can only access a noisy sample of the value of each item. These samples are normally distributed, with parameters such that their unbounded accumulation over one second is also normally distributed with a mean equal to kvs, where vs is the explicit value reported during the Ratings phase and k is a measure of signal-to-noise, and a standard deviation equal to 1.

Crucially, for each item, the noise in successive samples is correlated. To generate the correlated samples, we sample from a Markov chain using the Metropolis-Hastings algorithm (Chib and Greenberg, 1995). The target distribution is the normally distributed value function described in the previous paragraph. The proposal density is also normally distributed. Its width determines the degree of correlation between consecutive samples. Typically, the correlation between successive samples is considered a limitation of the Metropolis-Hastings algorithm. Here, however, it allows us to generate correlated samples from a target distribution. The standard deviation of the proposal density is . Higher values of τ result in a narrower proposal density, hence more strongly correlated samples. We sample from the same Markov chain across different trials in which the same item is presented, so that the last sample obtained about an item in a given trial is the initial state of the Markov chain the next time the item is presented.

At each moment (dt = 40ms), we sample one value for the left item and another for the right item, compute their difference (right minus left), and accumulate this difference until it crosses a threshold at +B0, signaling a rightward choice, or at −B0, signaling a leftward choice. The decision time is added to the non-decision time, μnd, to obtain the response time.

We fit the model to the data as follows. For each item, we simulate many Markov chains. In each trial, i, we take samples from each chain until the accumulation of these samples reaches one of the two decision thresholds. Then we calculate the likelihood (L) of obtaining the choice and the RT displayed by the participant on that trial as: where N = 1, 000 is the number of Markov chains, 𝟙 is an indicator function that takes the value 1 if the choice made on chain j is the same as the choice made by the participant on trial i and 0 otherwise, 𝒩 (x | y, z) is the normal probability density function with mean y and standard deviation z evaluated at x, and σnd is a parameter fit to the data.

When an item is presented again in a future trial, the initial state of each Markov chain depends on the state it was in the last time the item was presented. The initial state of each chain is obtained by sampling 1,000 values (one per chain) from the distribution given by the final state of each chain. The sampling is weighted by the value of Lj of each chain (Eq. 16), so that chains that better explained the choice and RT in the last trial are more likely to be sampled from in future trials.

The model has 5 parameters per participant: {k, B0, τ, μnd, σnd }, which were fit to maximize the sum, across trials, of the log of L using BADS (Acerbi and Ma, 2017).

The correlations in Fig. 11B were generated using the best-fitting parameters for each participant to simulate 100,000 Markov chains. We generate Markov chain samples independently for the left and right items over a 1-second period. To illustrate noise correlations, the simulations assume that the static value of both the left and right items is zero. We then calculate the difference in dynamic value (x) between the left and right items at each time (t) and for each of the Markov chains (i). Pearson’s correlation is computed between these differences at time zero, xi(t=0), and at time xi(t=τ), for different time lags τ. Correlations were calculated independently for each participant. Each trace in Fig. 11B represents a different participant.

fMRI analysis

Acquisition

Imaging data were acquired on a 3T GE MR750 MRI scanner with a 32-channel head coil. Functional data were acquired using a T2*-weighted echo planar imaging sequence (repetition time (TR) = 2 s, echo time (TE) = 22 ms, flip angle (FA) = 70°, field of view (FOV) = 192 mm, acquisition matrix of 96 × 96). Forty oblique axial slices were acquired with a 2 mm in-plane resolution positioned along the anterior commissure-posterior commissure line and spaced 3 mm to achieve full brain coverage. Slices were acquired in an interleaved fashion. We acquired three runs of the food choice task, each composed of 70 trials. Each of the food choice task functional runs consisted of 212 volumes and lasted 7 minutes. In addition to functional data, a single three-dimensional high-resolution (1 mm isotropic) T1-weighted full-brain image was acquired using a BRAVO pulse sequence for brain masking and image registration.

Preprocessing

Raw DICOM files were converted into Nifti file format and organized in the Brain Imaging Data Structure (BIDS) using dcm2niix (Li et al., 2016). Results included in this manuscript come from preprocessing performed using fMRIPrep 22.1.1 (Esteban et al. (2018b); Esteban et al. (2018a); RRID:SCR_016216), which is based on Nipype 1.8.5 (Gorgolewski et al. (2011); Gorgolewski et al. (2018); RRID:SCR_002502).

Anatomical data preprocessing

The T1-weighted (T1w) image was corrected for intensity non-uniformity (INU) with N4BiasFieldCorrection (Tustison et al., 2010), distributed with ANTs 2.3.3 (Avants et al., 2008, RRID:SCR_004757), and used as T1w-reference throughout the workflow. The T1w-reference was then skull-stripped with a Nipype implementation of the antsBrainExtraction.sh workflow (from ANTs), using OASIS30ANTs as target template. Volume-based spatial normalization to one standard space (MNI152NLin2009cAsym) was performed through nonlinear registration with antsRegistration (ANTs 2.3.3), using brain-extracted versions of both T1w reference and the T1w template. The following template was selected for spatial normalization: ICBM 152 Nonlinear Asymmetrical template version 2009c [Fonov et al. (2009), RRID:SCR_008796; TemplateFlow ID: MNI152NLin2009cAsym].

Functional data preprocessing

For each of the 3 BOLD runs per subject, the following preprocessing was performed. First, a reference volume and its skull-stripped version were generated using a custom methodology of fMRIPrep. Head-motion parameters with respect to the BOLD reference (transformation matrices, and six corresponding rotation and translation parameters) are estimated before any spatiotemporal filtering using mcflirt (FSL 6.0.5.1:57b01774, Jenkinson et al., 2002). The BOLD time-series (including slice-timing correction when applied) were resampled onto their original, native space by applying the transforms to correct for head-motion. These resampled BOLD time-series will be referred to as preprocessed BOLD in original space, or just preprocessed BOLD. The BOLD reference was then co-registered to the T1w reference using mri_coreg (FreeSurfer) followed by flirt (FSL 6.0.5.1:57b01774, Jenkinson and Smith, 2001) with the boundary-based registration (Greve and Fischl, 2009) cost-function. Co-registration was configured with six degrees of freedom. Several confounding time-series were calculated based on the preprocessed BOLD: framewise displacement (FD) and DVARS. FD was computed using two formulations following Power (absolute sum of relative motions, Power et al. (2014)) and Jenkinson (relative root mean square displacement between affines, Jenkinson et al. (2002)). FD and DVARS are calculated for each functional run, both using their implementations in Nipype (following the definitions by Power et al., 2014). The head-motion estimates calculated in the correction step were also placed within the corresponding confounds file. The confound time series were derived from head motion estimates (Satterthwaite et al., 2013). Frames that exceeded a threshold of 0.5 mm FD or 1.5 standardized DVARS were annotated as motion outliers. The BOLD time-series were resampled into standard space, generating a preprocessed BOLD run in MNI152NLin2009cAsym space. First, a reference volume and its skull-stripped version were generated using a custom methodology of fMRIPrep. All resamplings can be performed with a single interpolation step by composing all the pertinent transformations (i.e. head-motion transform matrices, susceptibility distortion correction when available, and co-registrations to anatomical and output spaces). Gridded (volumetric) resamplings were performed using antsApplyTransforms (ANTs), configured with Lanczos interpolation to minimize the smoothing effects of other kernels (Lanczos, 1964).

Many internal operations of fMRIPrep use Nilearn 0.9.1 (Abraham et al., 2014, RRID:SCR_001362), mostly within the functional processing workflow. For more details of the pipeline, see the section corresponding to workflows in fMRIPrep’s documentation.

Analysis

We conducted a GLM analysis to look at BOLD activity related to d-values, s-values, and the difference between the two. We ran four separate models.

Main fMRI Model included five regressors: (i) onsets for all valid trials, modeled with a duration equal to the average RT across all valid choices and participants; (ii) same onsets and duration as (i) modulated by RT demeaned across these trials within each run for each participant; (iii) same onsets and duration as (i) but modulated by the s-value of the chosen item demeaned across trials within each run for each participant; (iv) same onsets and duration as (i) but modulated by the d-value of the chosen item demeaned across these trials within each run for each participant; (v) onsets for missed trials. The map in Fig. 9 was generated using this model.

fMRI Model of s-value only included four regressors; all but regressor (iv) in Main fMRI Model. The map in Figure 9–Figure Supplement 1 top was generated using this model.

fMRI model of d-value only included four regressors; all but regressor (iii) in Main fMRI Model. The map in Figure 9–Figure Supplement 1 middle was generated using this model.

fMRI model of d-value − s-value only included four regressors; regressors (i) and (ii) were the same as in Main fMRI Model, regressor (iii) had the same onsets and duration as (i) but modulated by (d-value − s-value) of the chosen item demeaned across trials within each run for each participant, and regressor (iv) included onsets for missed trials. The map in Figure 9–Figure Supplement 1 bottom was generated using this model.

All four models included the six x, y, z translation and rotation motion parameters, FD, DVARS, and motion outliers obtained from textitfmriprep (described above) as confound regressors of no interest. All regressors were entered at the first level of analysis, and all (except the added confound regressors) were convolved with a canonical double-gamma hemodynamic response function. The time derivative of each regressor (except the added confounding regressors) was included in the model. No orthogonalization between regressors was performed. Models were estimated separately for each participant and run.

GLMs were estimated using FSL’s FMRI Expert Analysis Tool (FEAT). The first-level time-series GLM analysis was performed for each run per participant using FSL’s FILM. The first-level contrast images were then combined across runs per participant using fixed effects. The group-level analysis was performed using FMRIB’s Local Analysis of Mixed Effects (FLAME1) (Beckmann et al., 2003). Group-level maps were corrected to control the family-wise error rate using cluster-based Gaussian random field correction for multiple comparisons, with an uncorrected cluster-forming threshold of z=3.1 and corrected extent threshold of p < 0.05.